### 2015-03-20 cug_paper/01_introduction.tex
Add directory to hold the CUG paper

\section{Introduction}
\label{sec:intro}

Discuss the general idea and the rationale behind it

### 2015-03-20 cug_paper/02_design.tex
Add directory to hold the CUG paper

\section{Design And Implementation}
\label{sec:design}

Discuss how the software works.  Things like using CCI for small messages, bulk transfers happening by way of cudaMemcpy().  Mention Cray-specific stuff such as setting 'feature=gpudefault' in the qsub script.

### 2015-03-20 cug_paper/03_analysis.tex
Add directory to hold the CUG paper

\section{Analysis}
\label{sec:analysis}

\subsection{Experiment Setup}
Discuss what we ran, the parameters we swept and why we chose those parameters

\subsection{Results}
Discuss the results

### 2015-03-20 cug_paper/04_futurework.tex
Add directory to hold the CUG paper

\section{Future Work}
\label{sec:future}

Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use)

### 2015-03-20 cug_paper/05_conclusion.tex
Add directory to hold the CUG paper

\section{Conclusion}
\label{sec:conclusion}

Discuss what we learned; summarize

### 2015-03-20 client_src/new_test.cpp
Pass file offset into writeRemote()
writeRemote() needs to know the offset into the file where the data is
supposed to be written

@@ -150,7 +150,8 @@ int main( int argc, char **argv)
                 size_t bytesWritten;
                 while (totalBytesWritten < len) {
                     size_t bytesToWrite = len - totalBytesWritten;
-                    writeRemote( &buf[totalBytesWritten], bytesToWrite, &bytesWritten);
+                    writeRemote( &buf[totalBytesWritten], bytesToWrite,
+                                 totalBytesWritten, &bytesWritten);
                     if (bytesWritten == 0) {
                         // Out of space in the GPU memory - sleep briefly to
                         // keep from thrashing the system with write requests

### 2015-03-20 client_src/utils.cpp

@@ -305,7 +305,7 @@ int writeLocal(void *buf, streamsize len, ofstream &outf)
 // on what the daemon tells us).
 // returns a CCI_STATUS.  Number of bytes written (which may be less
 // than len, will be returned in bytesWritten)
-int writeRemote(void *buf, size_t len, size_t *bytesWritten)
+int writeRemote(void *buf, size_t len, size_t offset, size_t *bytesWritten)
 {   
 
     int ret = 0;
@@ -315,7 +315,7 @@ int writeRemote(void *buf, size_t len, size_t *bytesWritten)
     IoMsg sndMsg, replyMsg;
     sndMsg.writeRequest.type = WRITE_REQ;
     sndMsg.writeRequest.len = len;
-    sndMsg.writeRequest.offset = 0;
+    sndMsg.writeRequest.offset = offset;
        
     ret = cci_send( connection, &sndMsg, sizeof( sndMsg.writeRequest), &sndMsg, CCI_FLAG_NO_COPY);
     // Note: using the address of the buffer as the context...

### 2015-03-20 client_src/utils.h

@@ -31,7 +31,7 @@ int writeLocal(void *buf, std::streamsize len, std::ofstream &outf);
 // Write to the remote daemon (either the GPU or system ram depending
 // on what the daemon tells us).
 // returns a CCI_STATUS
-int writeRemote(void *buf, size_t len, size_t *bytesWritten);
+int writeRemote(void *buf, size_t len, size_t offset, size_t *bytesWritten);
 
 // Get the cache usage data from the daemon
 // returns a CCI_STATUS.  

### 2015-03-21 daemon_src/gpumem.cpp
Remove overly verbose debug message

@@ -29,8 +29,9 @@ uint64_t allocate_fcfs( void **devPtr, uint64_t reqLen)
 //             << "MB.  (" << (freeMem / (1024*1024)) << "MB free.)" << endl;
     
     if (freeMem < INACCESSIBLE_GPU_MEM) {
-        cerr << __func__ << ": Not enough free GPU mem to bother with a "
-             << "malloc attempt."  << endl;
+// Commented out message - too much output
+//        cerr << __func__ << ": Not enough free GPU mem to bother with a "
+//             << "malloc attempt."  << endl;
              return 0;
     }
     freeMem -= INACCESSIBLE_GPU_MEM;

### 2015-03-21 daemon_src/daemoncmdlineopts.cpp
Add help message for the -P option

@@ -98,6 +98,8 @@ void printUsage(char *name)
     cerr << "\t-G\tAmount of GPU ram to use for cache (in MB)." << endl
          << "\t\t0 for no GPU ram cache, -1 for unlimited. Default: "
          << MAX_GPU_RAM << endl;
+    cerr << "\t-P\tAttempt to pin threads to specific odd-numbered cores"
+         << endl;
     cerr << "\t-T\tNumber of background write threads. Default: " << WRITE_THREADS << endl;
     cerr << "\t-A\tSelect allocator for GPU memory.  Valid options are:" << endl;
     cerr << "\t\t1 - First come, first served allocator" << endl;

### 2015-03-24 client_src/utils.cpp
Ignore SIGCHLD signals before calling waitpid()
We're still getting odd SIGCHLD signals when we run on Titan.  This
change will at least mean they don't cause us to abort.  (I'd still like
to know why we're getting them, though.)

@@ -261,7 +261,7 @@ int finalizeIo(void)
     // dirty cache to disk) - wait for it...
     if (daemonPid != -1) {
         cerr << "Client: about to call waitpid() on PID " << daemonPid << endl;
-        //signal(SIGCHLD, SIG_IGN);
+        signal(SIGCHLD, SIG_IGN);
         //kill(daemonPid, SIGKILL);
         waitpid(daemonPid, NULL, 0);
         

### 2015-03-24 daemon_src/gpumem.cpp
Be a little more robust when allocating GPU memory
Even though we check before attempting to allocate memory, we still
occasionally get out-of-memory errors when calling cudaMalloc.  In such
a case, we don't need to abort the program.  We just return 0 bytes to
the calling function.

@@ -97,7 +97,18 @@ uint64_t allocate_block( void **devPtr, uint64_t reqLen)
         
     uint64_t attemptedLen = (reqLen < BLOCK_SIZE) ? reqLen : BLOCK_SIZE;
     
-    CUDA_CHECK_RETURN( cudaMalloc( devPtr, attemptedLen));
+    cudaError_t cudaErr = cudaMalloc( devPtr, attemptedLen);
+    if ( cudaErr != cudaSuccess) {
+        cerr << __func__ << ":" << __LINE__ 
+             << " cudaMalloc() failed - " <<  cudaGetErrorString( cudaErr)
+             << endl
+             << "Returning 0 and continuing." << endl
+             << "NOTE: This shouldn't happen! Potential out-of-memory "
+             << "conditions should have been caught before calling "
+             << "cudaMalloc()!" << endl;       
+        return 0;
+    }
+    
     return attemptedLen;
 }
 

### 2015-03-24 analysis/client_side.py
First cut at some analysis scripts
Using python to parse the result files and load them into a sqlite db.
From there, using MatPlotLib to draw plots of various data that's been
queried from the DB.

Can currently plot (per-rank or per-node) bandwidth vs. write size.

@@ -1,0 +1,152 @@
+#!/bin/env python
+#
+# A script for analyzing the client-side timing data from the io-agg
+# 'new_test' executable
+#
+# Sample output from a single rank (from a small run):
+#
+# len 134217728 start 1426876784488692 end 1426876784674608 elapsed 185916 MB/s 688.483
+# len 134217728 start 1426876796307678 end 1426876796485084 elapsed 177406 MB/s 721.509
+# len 134217728 start 1426876805916297 end 1426876806085703 elapsed 169406 MB/s 755.581
+# len 268435456 start 1426876818203246 end 1426876818585501 elapsed 382255 MB/s 669.71
+# len 268435456 start 1426876969876948 end 1426876970275177 elapsed 398229 MB/s 642.846
+# len 268435456 start 1426877063784120 end 1426877064170422 elapsed 386302 MB/s 662.694
+# len 536870912 start 1426877086479164 end 1426877087494383 elapsed 1015219 MB/s 504.325
+# len 536870912 start 1426877129710784 end 1426877130458838 elapsed 748054 MB/s 684.443
+# len 536870912 start 1426877166373242 end 1426877167339400 elapsed 966158 MB/s 529.934
+#
+#
+
+import os
+import os.path
+import sqlite3
+
+from optparse import OptionParser
+
+import plot_utils as pu
+
+#Plot function
+
+def timelines(y, xstart, xstop, color='b'):
+    """Plot timelines at y from xstart to xstop with given color."""   
+    plt.hlines(y, xstart, xstop, color, lw=4)
+    plt.vlines(xstart, y+0.03, y-0.03, color, lw=2)
+    plt.vlines(xstop, y+2, y-2, color, lw=2)
+
+
+TABLE_NAME = 'Results'
+
+def init_db( db_filename):
+    '''Initialize the database
+    
+    Opens the specified file, drops any existing tables and creates new ones.
+    '''
+    conn = sqlite3.connect(db_filename)
+    c = conn.cursor()
+    
+    c.execute( "DROP TABLE IF EXISTS %s;"%TABLE_NAME)
+    
+    # Create table
+    stmt = "CREATE TABLE %s ("                                \
+        "rank int, iteration int, node int, length int, "     \
+        "start int, end int, elapsed float, bw int);" % TABLE_NAME
+    c.execute(stmt)
+    conn.commit()
+    
+    return conn
+    
+def import_results( filename, rank, ranks_per_node, conn):
[…]

### 2015-03-24 analysis/plot_utils.py

@@ -1,0 +1,216 @@
+# Utility functions for making the plots we want
+# (Should probably be a python package...)
+
+import matplotlib.pyplot as plt
+import sqlite3
+import numpy as np
+
+#Plot function
+
+def timelines(y, xstart, xstop, color='b'):
+    """Plot timelines at y from xstart to xstop with given color."""   
+    plt.hlines(y, xstart, xstop, color, lw=4)
+    plt.vlines(xstart, y+0.03, y-0.03, color, lw=2)
+    plt.vlines(xstop, y+0.03, y-0.03, color, lw=2)
+
+
+
+
+
+TABLE_NAME="Results"
+
+# Plot average per-rank bandwidth as a function of write size
+# Assumes:
+#  First n/2 ranks used FCFS allocator, second n/2 ranks used balanced allocator
+def plot_per_rank_bw( conn, num_write_threads):
+    
+    c = conn.cursor()
+    c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
+    max_rank = c.fetchall()[0][0]
+    
+    # Queries that we'll execute (the plot will show 2 groupings)
+    statements = ["SELECT length, bw from %s WHERE rank < %d ORDER BY length;" \
+                  %(TABLE_NAME, max_rank/2),
+                  "SELECT length, bw from %s WHERE rank >= %d ORDER BY length;" \
+                  %(TABLE_NAME, max_rank/2) ]
+    
+    rects = [] # the Rectangle instances returned by calls to bar() below
+    fig, ax = plt.subplots() # initialize the plot
+    width = 0.35                # the width of the bars
+    colors = ['r','b']   # colors of the bars for the different groups
+    
+    for i in range( len(statements)):
+        c.execute( statements[i])
+        data=c.fetchall()
+        
+        prev_len = 0
+        length_dict = { }  # keys are the write lengths, values are numpy
+                        # arrays of the BW values
+        for row in data:
+            if row[0] != prev_len:
+                length_dict[row[0]] = []
+                prev_len = row[0]
+            length_dict[row[0]].append( row[1])
+            # NumPy arrays don't have append(), so start with a list
+            # (we'll convert to an array later)
+        
+        for k in length_dict.keys():
+            length_dict[k] = np.array(length_dict[k], float)
+        
+        # now have a dictionary of numpy arrays
+        # convert values from bytes/sec to MB/s and keys from bytes
+        # to MB
+        MB=1024*1024
+        for k in length_dict.keys():
[…]

### 2015-03-24 analysis/plot_utils.py
Add error bars showing min/max to the bar charts

@@ -79,19 +79,30 @@ def plot_per_rank_bw( conn, num_write_threads):
             averages.append( np.average( length_dict[k]))
         
         
+        # I want to display min & max values as error bars on
+        # each main bar.  MatPlotLib can do this, but it treats
+        # the upper & lower error values as values to be added
+        # or subtracted from the main value.  Hence the next two
+        # lines
+        low_err = np.array(averages) - np.array(mins)
+        high_err = np.array(maxes) - np.array(averages)
+        
         # debugging:
         print "Mins:", mins
         print "Maxes:", maxes
         print "Averages:", averages
+        print "Low error:", low_err
+        print "High error:", high_err
         
         # plot the data
         ind = np.arange(len(mins))  # the x locations for the groups
-        rects.append( ax.bar(ind + (width*i), averages, width, color=colors[i]))
+        rects.append( ax.bar(ind + (width*i), averages, width,
+                             color=colors[i], yerr=[low_err,high_err]))
     
     
     # add some text for labels, title and axes ticks
     ax.set_ylabel('Bandwidth (MB/s)')
-    ax.set_title('Per-Rank Bandwidth vs. Write Size')
+    ax.set_title('Average Per-Rank Bandwidth vs. Write Size')
     ax.set_xticks(ind+width)
     ax.set_xticklabels( x_vals)
     ax.set_xlabel( 'Write size (MB)')
@@ -177,19 +188,30 @@ def plot_per_node_bw( conn, num_write_threads):
             averages.append( np.average( length_dict[k]))
         
         
+        # I want to display min & max values as error bars on
+        # each main bar.  MatPlotLib can do this, but it treats
+        # the upper & lower error values as values to be added
+        # or subtracted from the main value.  Hence the next two
+        # lines
+        low_err = np.array(averages) - np.array(mins)
+        high_err = np.array(maxes) - np.array(averages)
+        
         # debugging:
         print "Mins:", mins
         print "Maxes:", maxes
         print "Averages:", averages
+        print "Low error:", low_err
+        print "High error:", high_err
         
         # plot the data
         ind = np.arange(len(mins))  # the x locations for the groups
-        rects.append( ax.bar(ind + (width*i), averages, width, color=colors[i]))
+        rects.append( ax.bar(ind + (width*i), averages, width,
+                             color=colors[i], yerr=[low_err,high_err]))
     
     
[…]

### 2015-03-24 analysis/client_side.py
Cleaning up analysis code
Re-factored the plotting code so that there was much less duplicated
code.  Also added a single-group per-node BW plot (for the jobs that
didn't use the GPU cache and thus don't have multiple allocator
functions to show).

@@ -144,7 +144,8 @@ def main():
         conn = sqlite3.connect( options.db_file)
         # connect to the existing db
               
-    
+
+    pu.plot_per_node_bw_non_gpu( conn, 2)
     pu.plot_per_rank_bw( conn, 2)
     pu.plot_per_node_bw( conn, 2)
 

### 2015-03-24 analysis/plot_utils.py

@@ -34,98 +34,13 @@ def plot_per_rank_bw( conn, num_write_threads):
                   "SELECT length, bw from %s WHERE rank >= %d ORDER BY length;" \
                   %(TABLE_NAME, max_rank/2) ]
     
-    rects = [] # the Rectangle instances returned by calls to bar() below
-    fig, ax = plt.subplots() # initialize the plot
-    width = 0.35                # the width of the bars
-    colors = ['r','b']   # colors of the bars for the different groups
-    
-    for i in range( len(statements)):
-        c.execute( statements[i])
-        data=c.fetchall()
-        
-        prev_len = 0
-        length_dict = { }  # keys are the write lengths, values are numpy
-                        # arrays of the BW values
-        for row in data:
-            if row[0] != prev_len:
-                length_dict[row[0]] = []
-                prev_len = row[0]
-            length_dict[row[0]].append( row[1])
-            # NumPy arrays don't have append(), so start with a list
-            # (we'll convert to an array later)
-        
-        for k in length_dict.keys():
-            length_dict[k] = np.array(length_dict[k], float)
-        
-        # now have a dictionary of numpy arrays
-        # convert values from bytes/sec to MB/s and keys from bytes
-        # to MB
-        MB=1024*1024
-        for k in length_dict.keys():
-            length_dict[k/MB] = length_dict[k] / MB
-            length_dict.pop(k)
-            
-        # Get the min, max & average values
-        # Note: we can't count on the order of items in a
-        # dictionary, so we're using lists
-        x_vals = length_dict.keys()
-        x_vals.sort()
-        mins = []
-        maxes = []
-        averages = []
-        for k in x_vals:
-            mins.append( np.amin( length_dict[k]))
-            maxes.append( np.amax( length_dict[k]))
-            averages.append( np.average( length_dict[k]))
-        
-        
-        # I want to display min & max values as error bars on
-        # each main bar.  MatPlotLib can do this, but it treats
-        # the upper & lower error values as values to be added
-        # or subtracted from the main value.  Hence the next two
-        # lines
-        low_err = np.array(averages) - np.array(mins)
-        high_err = np.array(maxes) - np.array(averages)
-        
-        # debugging:
-        print "Mins:", mins
-        print "Maxes:", maxes
[…]

### 2015-03-25 client_src/cmdlineopts.cpp
Partial support for blocking mode
Added code to enable blocking mode on the daemon.  Also have partial
support for blocking mode on the client.  (There's still a little more
work to do to enable it.)

THIS CODE IS UNTESTED!!  The version of CCI that I have doesn't support
blocking mode for the shared memory transport.

@@ -17,7 +17,6 @@ static struct option long_options[] = {
     {"min_len",          required_argument, 0, 'm'},
     {"max_len",          required_argument, 0, 'M'},
     {"client_blocking",  no_argument,       0, 'b'},
-    {"daemon_blocking",  no_argument,       0, 'B'},
     {"null_io",          no_argument,       0, 'n'},
     {"extra_ram",        required_argument, 0, 'e'},
     {"rma_buf",          required_argument, 0, 'r'},
@@ -33,7 +32,7 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
     {
         /* getopt_long stores the option index here. */
         int option_index = 0;
-        int c = getopt_long( argc, argv, "i:s:m:M:bBne:r:dD:N", long_options, &option_index);
+        int c = getopt_long( argc, argv, "i:s:m:M:bne:r:dD:N", long_options, &option_index);
 
         /* Detect the end of the options. */
         if (c == -1)
@@ -56,9 +55,6 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
                 break;
             case 'b':
                 opts.clientBlocking = true;
-                break;
-            case 'B':
-                opts.daemonBlocking = true;
                 break;
             case 'n':
                 opts.nullIo = true;
@@ -116,7 +112,6 @@ void printUsage(char *name)
     cerr << "\t-N\tDisable automatic startup of the remote daemon" << endl
          << "\t\t(Useful if running the daemon in the debugger.)" << endl;
     cerr << "** NOT IMPLEMENTED **" << "\t-b\tUse CCI blocking mode on client" << endl;
-    cerr << "** NOT IMPLEMENTED **" << "\t-B\tUse CCI blocking mode on iod daemon" << endl;
     cerr << "** NOT IMPLEMENTED **" << "\t-N\tUse NULL IO on iod daemon" << endl;
     cerr << endl;
     cerr << "A note about passing parameters to the daemon:  be sure to include the " << endl
@@ -131,7 +126,7 @@ void printUsage(char *name)
 CommandLineOptions::CommandLineOptions() :
      iters(ITERS), sleepSecs(SLEEP_SECS), minLen(MIN_LENGTH), maxLen(MAX_LENGTH),
[…]

### 2015-03-25 client_src/cmdlineopts.h

@@ -17,7 +17,6 @@
 #define EXTRA_RAM        0           // in MB 
 #define NULL_IO          false       // use null io option
 #define CLIENT_BLOCKING  false       // use CCI blocking mode on the client
-#define DAEMON_BLOCKING  false       // use CCI blocking mode on the server
 #define USE_DAEMON       false       // whether or not to hand writes off to the daemon process
 #define DAEMON_AUTOSTART true        // automatically start the daemon
 struct CommandLineOptions {
@@ -29,7 +28,6 @@ struct CommandLineOptions {
     int extraRam;  // in megabytes
     bool nullIo;
     bool clientBlocking;
-    bool daemonBlocking;
     size_t rmaBuf;  // in megabytes
     bool useDaemon;
     bool daemonAutostart;

### 2015-03-25 client_src/utils.cpp

@@ -29,8 +29,6 @@ static void handleSigchld( int sig);
 // variables that will need to be shared by multiple functions in this .cpp file
 pid_t daemonPid = -1;  // Process ID of the daemon
 cci_endpoint_t *endpoint = NULL;
-cci_os_handle_t *endpointFd = NULL; // file descriptor that can block waiting for
-                                    // progress on the endpoint
 cci_connection_t *connection = NULL;
 cci_rma_handle_t *local = NULL;
      
@@ -69,7 +67,7 @@ int startOneDaemon( char * const * daemonArgs)
 
 // Start up the daemon and set up the CCI connection
 // Returns a cci_status value, or a negated errno value (ie: -22 for EINVAL)
-int initIo(void *buffer, uint64_t len, uint32_t rank)
+int initIo(void *buffer, uint64_t len, uint32_t rank, cci_os_handle_t *endpointFd)
 {
     int ret = CCI_SUCCESS;
     uint32_t caps = 0;

### 2015-03-25 client_src/utils.h

@@ -4,6 +4,7 @@
 #ifndef _UTILS_H_
 #define _UTILS_H_
 
+#include <cci.h>
 #include <string.h>
 #include <stdint.h>
 
@@ -17,7 +18,9 @@ int startOneDaemon( char * const * daemonArgs);
 
 // Set up the CCI connection
 // Returns a cci_status value, or a negated errno value (ie: -22 for EINVAL)
-int initIo( void *buffer, uint64_t len, uint32_t rank);
+// If endpointFD is non-NULL, the function will return file descriptor that can
+//  block waiting for progress on the endpoint
+int initIo( void *buffer, uint64_t len, uint32_t rank, cci_os_handle_t *endpointFd=NULL); 
 
 // Shut down the CCI connection
 // Note: doesn't touch the daemon.  (Presumably, it will shut itself down.)

### 2015-03-25 daemon_src/daemon.cpp

@@ -12,6 +12,7 @@
 
 #include <assert.h>
 #include <cci.h>
+#include <poll.h>
 #include <pthread.h>
 #include <semaphore.h>
 #include <string.h>
@@ -49,7 +50,7 @@ static int cciSetup( char **uri);
 static void pinToCore( int core);
 
 // Handle all the CCI events
-static void commLoop();
+static void commLoop( int fd=-1);
 
 // thread function for copying data out of cache and writing it to disk
 void *writeThread( void *);
@@ -119,6 +120,12 @@ int main(int argc, char *argv[])
     
     gpuAlloc = cmdOpts.gpuAlloc;
     
+    //TODO: Clean this up.  Get rid of the global variable!
+    int eventBlockingFd;
+    if (cmdOpts.blockingMode) {
+        endpointFd = &eventBlockingFd;
+    }
+
     ret = cciSetup( &uri);
 
     memset(hostname, 0, sizeof(hostname));
@@ -677,12 +684,22 @@ static void handleSend( cci_event_t * event)
 
 
 // Handle all the CCI events
-static void commLoop()
+static void commLoop( int fd)
 {
     int ret = 0;
     cci_event_t *event = NULL;
-        
+    struct pollfd pfd;
+
+    if (fd > 0) {
+        // if we were passed a file descriptor, setup to poll it
+        pfd.fd = fd;
+        pfd.events = POLLIN;
+    }
+
     while (!shuttingDown) {
+        if (fd > 0) {
+            poll(&pfd, 1, 500); // wake up after 500ms (so we can check for shutdown)
+	}
         ret = cci_get_event(endpoint, &event);
         if (ret) { 
             if (ret != CCI_EAGAIN) {

### 2015-03-25 daemon_src/daemoncmdlineopts.cpp

@@ -16,6 +16,7 @@ static struct option long_options[] = {
     {"gpu_allocator",  required_argument, 0, 'A'},
     {"write_threads",  required_argument, 0, 'T'},
     {"core_pinning",   no_argument,       0, 'P'},
+    {"blocking_mode",  no_argument,       0, 'B'},
     {0, 0, 0, 0} };
     
     
@@ -36,7 +37,7 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
     {
         /* getopt_long stores the option index here. */
         int option_index = 0;
-        int c = getopt_long( argc, argv, "S:G:A:T:P", long_options, &option_index);
+        int c = getopt_long( argc, argv, "S:G:A:T:PB", long_options, &option_index);
 
         /* Detect the end of the options. */
         if (c == -1)
@@ -75,6 +76,9 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
             case 'P':
                 opts.corePinning = true;
                 break;
+            case 'B':
+                opts.blockingMode = true;
+                break;
                 
             default:  // unrecognized option
                 return false;
@@ -100,6 +104,8 @@ void printUsage(char *name)
          << MAX_GPU_RAM << endl;
     cerr << "\t-P\tAttempt to pin threads to specific odd-numbered cores"
          << endl;
+    cerr << "\t-B\tBlocking mode: the event loop blocks waiting for events, instead of polling" << endl;
+    cerr << "\tWARNING: Blocking mode is untested!  It requires an experimental version of CCI!" << endl;
     cerr << "\t-T\tNumber of background write threads. Default: " << WRITE_THREADS << endl;
     cerr << "\t-A\tSelect allocator for GPU memory.  Valid options are:" << endl;
     cerr << "\t\t1 - First come, first served allocator" << endl;

### 2015-03-25 daemon_src/daemoncmdlineopts.h

@@ -13,7 +13,7 @@
 #define DEFAULT_ALLOC    allocate_fcfs  // "First come / first served" allocation
 #define WRITE_THREADS    1 // one background write thread
 #define CORE_PINNING     false 
-// TODO: Do we need an option for blocking mode?
+#define BLOCKING_MODE    false
 
 struct CommandLineOptions {
     // cache sizes
@@ -25,10 +25,10 @@ struct CommandLineOptions {
                         // Pointer to GPU mem allocation functions
     
     bool corePinning;  // Attempt to pin threads to odd numbered cores?
-    
+    bool blockingMode; // Run the comm loop in blocking mode    
     CommandLineOptions() : maxSysRam(MAX_SYS_RAM), maxGpuRam(MAX_GPU_RAM),
                            writeThreads( WRITE_THREADS), gpuAlloc( DEFAULT_ALLOC),
-                           corePinning( CORE_PINNING) { }
+                           corePinning( CORE_PINNING), blockingMode(BLOCKING_MODE) { }
     
 };
 

### 2015-03-26 client_src/cmdlineopts.cpp
Check for unexpected positional options on the command line

@@ -84,6 +84,17 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
                 
         }
     } // end while(1)
+    
+    // Check for any 'extra' positional arguments (ie: ones that didn't
+    // start with '-' or '--'
+    if (optind != argc) {
+        cerr << "WARNING (" << argv[0] << "): Unexpected positional arguments: ";
+        for (int i=optind; i < argc; i++) {
+            cerr << argv[i] << " ";
+        }
+        cerr << endl;
+        cerr << "Did you forget a '-' character?" << endl;
+    }
 // TODO: We ought to add some code to validate the options we've got - there's a numer
 // of combinations that are invalid (ex: sleepSecs of -1 requires useDaemon)
     return true;  // no parse errors

### 2015-03-26 daemon_src/daemoncmdlineopts.cpp

@@ -85,6 +85,17 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
                 
         }
     } // end while(1)
+    
+    // Check for any 'extra' positional arguments (ie: ones that didn't
+    // start with '-' or '--'
+    if (optind != argc) {
+        cerr << "WARNING (" << argv[0] << "): Unexpected positional arguments: ";
+        for (int i=optind; i < argc; i++) {
+            cerr << argv[i] << " ";
+        }
+        cerr << endl;
+        cerr << "Did you forget a '-' character?" << endl;
+    }
         
     return true;  // no parse errors
 }

### 2015-03-26 client_src/cmdlineopts.cpp
Change how the -s parameter works
If -s is positive, then it's the same as before: sleep for a fixed
amount of time.  If -s is negative, then the code waits until the daemon
reports that the cache is empty.  It then calls sync() and waits the
specified amount of time (presumably to let the OS page cache drain).

@@ -110,7 +110,9 @@ void printUsage(char *name)
     cerr << "where:" << endl;
     cerr << "\t-i\tNumber of iterations (default " << ITERS << ")" << endl;
     cerr << "\t-s\tSeconds between writes (default " << SLEEP_SECS << ")" << endl;
-    cerr << "\t  \t  (-1 says to sleep until the daemon has flushed its cache.)" << endl;
+    cerr << "\t  \t  Positive numbers sleep for the specified amount of time." << endl;
+    cerr << "\t  \t  Negative numbers wait until the daemon says the cache is empty" << endl
+         << "\t  \t  and then sleep an additional specified number of seconds." << endl;
     cerr << "\t-m\tMinimun length (default " << MIN_LENGTH << ")" << endl;
     cerr << "\t-M\tMaximum length (default " << MAX_LENGTH << ")" << endl;
     

### 2015-03-26 client_src/new_test.cpp

@@ -179,18 +179,19 @@ int main( int argc, char **argv)
             // Wait for the cache to drain (by sleeping)
             // If we're in automatic mode, check with the daemon for the
             // cache usage.  Otherwise, sleep for a fixed amount of time
-            if (cmdOpts.sleepSecs == -1) {
+            if (cmdOpts.sleepSecs < 0) {
                 if (cmdOpts.useDaemon) {
                 // can't call checkCacheUsage() if there's no daemon
                     bool isEmpty = false;
                     checkCacheUsage( &isEmpty);
                     while ( ! isEmpty) {
-                        usleep( 100 * 1000);
+                        usleep( 10 * 1000);
                         checkCacheUsage( &isEmpty);
                     }
                 }
                 
                 sync();  // daemon's cache is empty - this flushes the OS pagecache
+                sleep( cmdOpts.sleepSecs * -1);  // give the OS pagecache time to drain
             } else if (cmdOpts.sleepSecs > 0) {
                 sleep( cmdOpts.sleepSecs);
             }

### 2015-03-26 client_src/cmdlineopts.cpp
Update the help text
The changes to the help text in the previous commit didn't mention
calling sync()

@@ -112,7 +112,8 @@ void printUsage(char *name)
     cerr << "\t-s\tSeconds between writes (default " << SLEEP_SECS << ")" << endl;
     cerr << "\t  \t  Positive numbers sleep for the specified amount of time." << endl;
     cerr << "\t  \t  Negative numbers wait until the daemon says the cache is empty" << endl
-         << "\t  \t  and then sleep an additional specified number of seconds." << endl;
+         << "\t  \t  and then call sync().  Once sync() returns, the client sleeps" << endl
+         << "\t  \t  an additional specified number of seconds." << endl;
     cerr << "\t-m\tMinimun length (default " << MIN_LENGTH << ")" << endl;
     cerr << "\t-M\tMaximum length (default " << MAX_LENGTH << ")" << endl;
     

### 2015-03-26 client_src/new_test.cpp
Print git hash & build date on startup
Borrowed this code from load_du_data: the Makefile checks the git hash
and includes it as a macro define on the command line.  The client
program prints this message along with the compile date at startup so we
have a record of exactly which executable ran.

@@ -50,8 +50,8 @@ int main( int argc, char **argv)
         printUsage(argv[0]);
         return -1;
     }
-
-    
+   
+   
     // Make sure the GPU is in the proper compute mode, or this isn't going to work
     struct cudaDeviceProp prop;
     CUDA_CHECK_RETURN( cudaGetDeviceProperties ( &prop, 0));
@@ -79,6 +79,20 @@ int main( int argc, char **argv)
     MPI_Init(&argc, &argv);
     MPI_Comm_rank(MPI_COMM_WORLD, &rank);
     MPI_Comm_size(MPI_COMM_WORLD, &ranks);
+   
+    if (rank == 0) {
+        // Print out an initial statement about compile time and git commit
+        // hash just to ensure that the executable we're running is the one
+        // we actually expect to run.
+        //
+        // NOTE: GIT_COMMIT is expected to be defined on the compiler command
+        // line.  Check the Makefile for details.
+#ifndef GIT_COMMIT
+#define GIT_COMMIT "UNDEFINED"
+#endif
+        cerr << argv[0] << ": Compiled on " << __DATE__ << " at " << __TIME__
+             << "   Git commit hash: " << GIT_COMMIT << endl;
+    }
     
     
     unsigned char* buf;

### 2015-03-27 client_src/cmdlineopts.cpp
Add option to no use so much memory
With the new option (-h) the client only allocates enough memory for
half of the largest write size.  The amount of data that's written stays
the same, and we just re-use the buffer as necessary.

Necessary to do 16 ranks per node at 2GB max size on Titan.

@@ -19,6 +19,7 @@ static struct option long_options[] = {
     {"client_blocking",  no_argument,       0, 'b'},
     {"null_io",          no_argument,       0, 'n'},
     {"extra_ram",        required_argument, 0, 'e'},
+    {"half_mem",         no_argument,       0, 'h'},
     {"rma_buf",          required_argument, 0, 'r'},
     {"use_daemon",       no_argument,       0, 'd'},
     {"daemon_arg",       required_argument, 0, 'D'},
@@ -32,7 +33,7 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
     {
         /* getopt_long stores the option index here. */
         int option_index = 0;
-        int c = getopt_long( argc, argv, "i:s:m:M:bne:r:dD:N", long_options, &option_index);
+        int c = getopt_long( argc, argv, "i:s:m:M:bne:hr:dD:N", long_options, &option_index);
 
         /* Detect the end of the options. */
         if (c == -1)
@@ -61,6 +62,9 @@ bool parseCmdLine( int argc, char **argv, CommandLineOptions &opts)
                 break;
             case 'e':
                 opts.extraRam = atoi(optarg);
+                break;
+            case 'h':
+                opts.halfMem = true;
                 break;
             case 'r':
                 opts.rmaBuf = atol(optarg);
@@ -120,7 +124,10 @@ void printUsage(char *name)
     cerr << "** NOT IMPLEMENTED **" << "\t-r\tSize of the RMA buffer (for aggregation). "
          << "In MB (default " << MAX_LENGTH / (1024*1024) << ")" << endl;
     cerr << "** NOT IMPLEMENTED **" << "\t-e\tAllocate extra memory. In MB (default " << EXTRA_RAM << ")" << endl;
-    
+    cerr << "\t-h\tOnly allocate half the memory needed for the largest write size (and therefore" << endl
+         << "\t\twrite the memory buffer twice for the largest writes).  Needed for 2GB write testing" << endl
+         << "\t\ton Titan." << endl;
+         
     cerr << "\t-d\tSend writes over to the remote daemon" << endl;
     cerr << "\t-D\tPass the argument through to the daemon" << endl;
     cerr << "\t-N\tDisable automatic startup of the remote daemon" << endl

### 2015-03-27 client_src/cmdlineopts.h

@@ -19,6 +19,7 @@
 #define CLIENT_BLOCKING  false       // use CCI blocking mode on the client
 #define USE_DAEMON       false       // whether or not to hand writes off to the daemon process
 #define DAEMON_AUTOSTART true        // automatically start the daemon
+#define HALF_MEM         false       // allocate half memory for a complete write request
 struct CommandLineOptions {
 
     unsigned iters;
@@ -29,6 +30,7 @@ struct CommandLineOptions {
     bool nullIo;
     bool clientBlocking;
     size_t rmaBuf;  // in megabytes
+    bool halfMem;
     bool useDaemon;
     bool daemonAutostart;
     char **daemonArgs;   

### 2015-03-27 client_src/new_test.cpp

@@ -34,6 +34,8 @@ struct TimeStamp {
 };
 
 
+static uint64_t MIN64( uint64_t a, uint64_t b) { return ((a<b)?a:b); }
+
 static void initBuffer(void *buf, size_t len, unsigned seed);
 
 int main( int argc, char **argv)
@@ -96,10 +98,14 @@ int main( int argc, char **argv)
     
     
     unsigned char* buf;
+    unsigned long bufLen = cmdOpts.maxLen;
+    if (cmdOpts.maxLen) {
+        bufLen /= 2;
+    }
     bool bufIsPinned = false;
     
 #ifndef DISABLE_CLIENT_PINNED_MEMORY
-    cudaError_t cudaErr = cudaMallocHost( &buf, cmdOpts.maxLen);
+    cudaError_t cudaErr = cudaMallocHost( &buf, bufLen);
     if (cudaErr == cudaSuccess) {
         bufIsPinned = true;
     } else {
@@ -112,7 +118,7 @@ int main( int argc, char **argv)
 
     if (! bufIsPinned) {
         // either the cudaMallocHost() call failed, or it wasn't compiled in
-        buf = new unsigned char[cmdOpts.maxLen];
+        buf = new unsigned char[bufLen];
     }
     
     // If we're not using the daemon, we'll need a file we can write to
@@ -130,7 +136,7 @@ int main( int argc, char **argv)
         
     // Initialize CCI
     if (cmdOpts.useDaemon) {
-        int rc = initIo(buf, cmdOpts.maxLen, rank);
+        int rc = initIo(buf, bufLen, rank);
         if (rc) {
             cerr << "CCI error " << rc << " during initialization: " 
                  << cci_strerror( NULL,(cci_status)rc) << endl;
@@ -152,7 +158,7 @@ int main( int argc, char **argv)
         for (unsigned i = 0; i < cmdOpts.iters; i++) {           
             TimeStamp ts( len);
 
-            initBuffer(buf, len, rank + i);
+            initBuffer(buf, bufLen, rank + i);
 
             /* Sync up */
             MPI_Barrier(MPI_COMM_WORLD);
@@ -164,7 +170,17 @@ int main( int argc, char **argv)
                 size_t bytesWritten;
                 while (totalBytesWritten < len) {
                     size_t bytesToWrite = len - totalBytesWritten;
-                    writeRemote( &buf[totalBytesWritten], bytesToWrite,
+                    // This math get's a little strange: We want to write out the buffer,
[…]

### 2015-03-27 client_src/utils.cpp

@@ -292,7 +292,6 @@ static void handleSigchld( int sig)
 // write to a local file
 int writeLocal(void *buf, streamsize len, ofstream &outf)
 {   
-    outf.seekp( 0);
     outf.write( (const char *)buf, len);
     return 0;
 }

### 2015-03-27 cug_paper/01_introduction.tex
Paper:  Lots of work on the CUG paper.

@@
 
-Discuss the general idea and the rationale behind it
+\todo[inline]{Discuss the general idea and the rationale behind it}
+
+Titan, the Cray XK7 at Oak Ridge National Laboratory, has 18,688 compute nodes.
+Each node consists of a 16-core AMD CPU, an NVIDIA GPU and 32GB ram.
+In addition, there is another 6GB of ram on each GPU card.
+Not surprisingly, there are applications that do not use all the resources on a compute node.
+
+Specifically, approximately one third of the applications that run on Titan do not use the GPU at all.\todo{Find a citation for this.}
+There may be good reasons for this.
+Utilizing the GPU's requires changes to the application and it's possible the developers haven't had the time to do that.
+It's also possible that the type of computations the application performs simply don't lend themselves to GPU-based acceleration.
+Regardless of why, it's clear that not all applications are benefiting from the available hardware.
+
+\todo[inline]{Abrupt transition: we go from talking about hardware to talking about software...}
+
+In most HPC applications, filesystem writes typically follow a very 'bursty' pattern.
+That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output when the application writes its checkpoint or output files.
+At ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it.
+This means that the performance that a user sees at any given time will likely vary dramatically.
+
+Having a sufficiently large write-back cache could improve applications' write performance in two ways.
+First, if the cache is large enough to hold the entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.
+It can write to the cache and then resume its computations while the cache drains to the filesystem.
+Secondly, write performance should be more predictable.
+From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and that is not effected by outside influences.
+
+ The GPU memory and integer cores can be used to implement such a cache without taking resources away from the main application.
+
+Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory.
+Its bandwidth is too low and its latency is too high.
+However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast.
+Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load.
+This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.
+
+\begin{comment}
+
+Specifically, though AMD markets the processor as having 16 cores, it actually has only 8 floating-point units.
+As such, some applications that are floating-point intensive run with only 8 processes per node.
+This actually leaves 8 cores available so long as they primarily execute integer instructions.
+Furthermore, some applications do not make use of the GPU and thus it, and the ram attached to it, also go unused.
+
+Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory.
+Its bandwidth is too low and its latency is too high.
+However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast.
+Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load.
+This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.
+
+In HPC applications, most filesystem writes follow a very ?bursty?
+pattern.
+That is, there will be long periods with no file output while the application is in a compute phase, followed by a short period of intense output when the application writes its checkpoint or output files.
+Furthermore, at ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it.
+This means that the performance that a user sees at any given time will likely vary dramatically.
+
+Having a sufficiently large write-back cache would alleviate both of these problems.
+Assuming the cache is large enough, then the 'bursty' nature of the output pattern can be smoothed out since there will be time for the traffic to drain from the cache before the next burst arrives.
+Equally importantly, write performance should be more predictable.
+From the application's point of view, write speed is only limited by how fast data can be copied into the cache, and that is not effected by outside influences.
+The GPU memory and integer cores can be used to implement such a cache without taking resources away from the main application.
+
+This paper demonstrates a technique for providing such a write-back cache to applications.
+We use a synthetic benchmark to demonstrate performance without such a cache, with a cache that utilizes regular system ram and with a cache that utilizes ram on the GPU card.
+Performance benefits and any potential trade-offs and caveats are discussed.
+\end{comment}

### 2015-03-27 cug_paper/02_design.tex
Paper:  Lots of work on the CUG paper.

@@
 
-Discuss how the software works.
+\todo[inline]{Discuss how the software works.
 Things like using CCI for small messages, bulk transfers happening by way of cudaMemcpy().
 Mention Cray-specific stuff such as setting 'feature=gpudefault' in the qsub script.
+}
+
+\subsection{Overview}
+\label{subsec:overview}
+This idea was tested using a synthetic benchmark application written by the authors.
+The benchmark consists of two executables: the main executable which simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
+The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.
+Note that the daemon does \emph{not} copy the data into the GPU memory.
+There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.
+\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.
+See section \ref{subsec:cray-specific}.}
+
+With the daemon \todo{Should 'daemon' be plural?} started, the application begins its main execution.  In this case, since it's just a synthetic 
+I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.
+For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.
+
+When it's time to write, the application sends a short message to the daemon.
+This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.
+For production code, something more flexible will obviously be needed.}
+\todo[inline]{We need a diagram showing the communication between the processes.}  The daemon will attempt to allocate memory on the GPU.
+The daemon will then reply to the application with the size of memory that was actually allocated and a memory handle that the application can use to access the GPU memory.
+Note that the size value may be less than what the application asked for.
+
+Upon receiving the reply, the application opens the CUDA memory handle and copies the specified amount of data up to the GPU memory.
+When the copy has completed, the application closes the memory handle and sends a message back to the daemon say that this particular write has completed.
+If the application has more data to write, it can send another write request, or else it can continue with its calculations.
+
+Note that the data has not made it out to the filesystem yet; it's just sitting in GPU memory.  Upon receiving the 'write done' message from the application, the daemon now knows it's free to
+
+The daemon maintains one or more background threads that are responsible for copying data out of GPU memory into normal system memory and then writing that data to the filesystem.
+
+The background write thread(s) operates asynchronously.
+
+Each thread copies data out of GPU memory and into a buffer in system memory.
+From there, it is written out to the filesystem.
+The system memory buffer is deliberation kept fairly small - 16MB - in order to conserve resources.
+Testing done on Titan shows that 16MB block sizes are large enough for cudaMemcpy() to give good performance.\cite{ac_guide}  Larger buffers would simply use up memory without improving performance noticeably.
+
+\todo[inline]{Do we need to get into the details of cuda IPC functions? cudaIpcGetMemHandle, cudaIpcOpenMemHandle, cudaIpcCloseMemHandle}
+
+From the application's point of view, the write has completed once the 'write done' message has been successfully sent.
+
+Once the applications sends the 'write done' message
+
+\todo[inline]{Need to mention CCI - short messages, low latency.  Citation reference is "atchley11:cci"}
+
+The GPU attempts to perform some load-balancing by restricting the amount of memory allocated for any single request.
+Currently this limit is 16MB.
+If an application wants to write 512MB, it will therefore have to make 32 separate requests.
+If other ranks on the node are trying to write at the same time, their requests will all be interleaved and hopefully rank will be starved.
+
+It's debatable whether this sort of load balancing is necessary or even desirable.
+Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).
+However, that same testing showed that one of the CUDA functions - cudaIpcOpenMemHandle() - became unreliable once GPU memory allocations exceeded 512MB.
+Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.
+
+\subsection{Cray-Specific Details}
+\label{subsec:cray-specific}
+
+\todo[inline]{Need an introduction paragraph for this subsection.}
+The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.
+On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.
+Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.
+This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called 'gpudefault'.
+The naming scheme comes from the CUDA documentation.
+On a standard CUDA install, the GPU would in fact default to allowing access from multiple processes.
+Titan is configured differently, however.
+Thus the need to explicitly tell qsub to switch to the 'default' mode.}
+
+\todo[inline] {talk about actually starting up the daemon - fork(), open(excl), exec() }

### 2015-03-27 cug_paper/03_analysis.tex
Paper:  Lots of work on the CUG paper.

@@
 \subsection{Experiment Setup}
-Discuss what we ran, the parameters we swept and why we chose those parameters
+\label{subsec:exp_setup}
+\todo[inline]{Discuss what we ran, the parameters we swept and why we chose those parameters}
+
+Two experiments
+First one: 8 ranks per node, 64 nodes.
+The application ranks were placed on the even numbered cores and the daemon threads were pinned to the odd numbered cores.
+Write sizes varied from 1MB to 2GB and the daemon was run with 1, 2 \& 4 background threads.
+
+Second experiment: 16 ranks per node, 64 nodes.
+The application ranks consumed all cores on the node and the daemon threads were left free to be scheduled on any core.
+The daemon was run with only a single write thread.
+(Results from the first experiment showed no improvement when using multiple write threads, and on this test the cores were already oversubscribed.
+Adding more write threads would only exacerbate that problem.)
+
+Plots show the average node write speeds (i.e.: the sum of the write speeds reported by each rank running on the node) as seen from the perspective of the application.
 
 \subsection{Results}
-Discuss the results
+\label{subsec:results}
+\todo[inline]{Discuss the results}

### 2015-03-27 cug_paper/04_futurework.tex
Paper:  Lots of work on the CUG paper.

@@
 
-Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use)
+\todo[inline]{Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use), measure/quantify any negative impacts on app performance (particularly for apps that run 16 ranks/node)}
+
+\todo[inline]{ Summit will have node-local SSD's and that a variant of this code could be quite useful for applications that haven't been modified to take advantage of them more directly.
+Not sure if we can mention Summit details yet.
+Not sure if we should mention Summit at a Cray conference...
+Maybe this paragraph should be in terms of generic burst-buffer support.}

### 2015-03-30 analysis/client_side.py
More updates to the analysis scripts
Added a couple more pre-packaged plot functions to plot_utils.py.  Also
commented out the sample plot calls in client_side.py

@@ -144,10 +144,10 @@ def main():
         conn = sqlite3.connect( options.db_file)
         # connect to the existing db
               
-
-    pu.plot_per_node_bw_non_gpu( conn, 2)
-    pu.plot_per_rank_bw( conn, 2)
-    pu.plot_per_node_bw( conn, 2)
+    # Some sample plot commands
+#    pu.plot_per_node_bw_non_gpu( conn, 2)
+#    pu.plot_per_rank_bw( conn, 2)
+#    pu.plot_per_node_bw( conn, 2)
 
 if __name__ == "__main__":
     main()

### 2015-03-30 analysis/plot_utils.py

@@ -78,11 +78,70 @@ def plot_per_node_bw_non_gpu( conn, num_write_threads):
                       x_label = 'Write size (MB)',
                       y_label = 'Bandwidth (MB/s)')
                      
+                     
+# Bar chart of the per-node BW vs. write size.  Assumes 192 nodes,
+# nodes 0-63: 1 write thread, nodes 64-127: 2 write threads and
+# nodes 128-191: 4 write threads
+def plot_per_node_bw_group_by_write_thread( conn):
+    c = conn.cursor()
+    #c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
+    #max_rank = c.fetchall()[0][0]
+    
+    # Queries that we'll execute (the plot will show 3 groupings)
+    statements = ["SELECT length, SUM(bw) from %s WHERE node < 64 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME,
+                  "SELECT length, SUM(bw) from %s WHERE rank >= 64 and rank < 128 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME,
+                  "SELECT length, SUM(bw) from %s WHERE rank < 192 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME]
+    # The GROUP BY & SUM() clauses cause us to sum the bw value for all rows
+    # with identical values for the (node, iteration) pair.  Since iteration
+    # increments on every write(), the only rows with identical (node,iteration)
+    # values will be the ones for different ranks.  (ie: no need to include
+    # length in the GROUP BY clause)
+    
+    generic_bar_plot( conn, statements, 
+                      title = 'Average Per-Node Bandwidth vs. Write Size',
+                      x_label = 'Write size (MB)',
+                      y_label = 'Bandwidth (MB/s)',
+                      legend=('1 Write Thread', '2 Write Threads',
+                              '4 Write Threads'))
+    
+
+
+# Bar chart of the per-RANK BW vs. write size.  Assumes 192 nodes,
+# nodes 0-63: 1 write thread, nodes 64-127: 2 write threads and
+# nodes 128-191: 4 write threads
+def plot_per_rank_bw_group_by_write_thread( conn):
+    c = conn.cursor()
+    #c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
+    #max_rank = c.fetchall()[0][0]
+    
+    # Queries that we'll execute (the plot will show 3 groupings)
[…]

### 2015-03-31 analysis/plot_utils.py
More work on the python analysis code
Fixed a couple of serious errors in the SQL statements that resulted in
extracting the wrong rows from the db table.

Added another plot type an da few cosmetic fixes

@@ -83,14 +83,14 @@ def plot_per_node_bw_non_gpu( conn, num_write_threads):
 # nodes 0-63: 1 write thread, nodes 64-127: 2 write threads and
 # nodes 128-191: 4 write threads
 def plot_per_node_bw_group_by_write_thread( conn):
-    c = conn.cursor()
+    #c = conn.cursor()
     #c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
     #max_rank = c.fetchall()[0][0]
     
     # Queries that we'll execute (the plot will show 3 groupings)
     statements = ["SELECT length, SUM(bw) from %s WHERE node < 64 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME,
-                  "SELECT length, SUM(bw) from %s WHERE rank >= 64 and rank < 128 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME,
-                  "SELECT length, SUM(bw) from %s WHERE rank < 192 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME]
+                  "SELECT length, SUM(bw) from %s WHERE node >= 64 and node < 128 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME,
+                  "SELECT length, SUM(bw) from %s WHERE node >= 128 and node < 192 GROUP BY node, iteration ORDER BY length;"%TABLE_NAME]
     # The GROUP BY & SUM() clauses cause us to sum the bw value for all rows
     # with identical values for the (node, iteration) pair.  Since iteration
     # increments on every write(), the only rows with identical (node,iteration)
@@ -110,14 +110,14 @@ def plot_per_node_bw_group_by_write_thread( conn):
 # nodes 0-63: 1 write thread, nodes 64-127: 2 write threads and
 # nodes 128-191: 4 write threads
 def plot_per_rank_bw_group_by_write_thread( conn):
-    c = conn.cursor()
+    #c = conn.cursor()
     #c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
     #max_rank = c.fetchall()[0][0]
     
     # Queries that we'll execute (the plot will show 3 groupings)
     statements = ["SELECT length, bw from %s WHERE node < 64 ORDER BY length;"%TABLE_NAME,
-                  "SELECT length, bw from %s WHERE rank >= 64 and rank < 128 ORDER BY length;"%TABLE_NAME,
[…]

### 2015-04-01 cug_paper/01_introduction.tex
Paper:  Introduction section has reached 'rough draft' state

@@
 
-\todo[inline]{Discuss the general idea and the rationale behind it}
+In most HPC applications, filesystem writes typically follow a very 'bursty' pattern.
+That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.
+Also, at ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it.
+This means that the performance that a user sees at any given time will likely vary dramatically.
+
+Having a sufficiently large node-local write-back cache could improve applications' performance in two ways.
+First, if the cache is large enough to hold an entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.
+It can write to the cache and then resume its computations while the cache drains to the filesystem.
+Secondly, write performance should be more predictable.
+From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and since the cache is local to the node, it is not effected by outside influences.
 
@@
 
-Specifically, approximately one third of the applications that run on Titan do not use the GPU at all.\todo{Find a citation for this.}
+To start with, during 2014, approximately 50\% of the compute time on Titan was used on applications that did not use the GPU at all.\cite{oareport}
 There may be good reasons for this.
@@
 It's also possible that the type of computations the application performs simply don't lend themselves to GPU-based acceleration.
-Regardless of why, it's clear that not all applications are benefiting from the available hardware.
+Regardless of why, it's clear that not all applications are benefiting from the GPU.
 
-\todo[inline]{Abrupt transition: we go from talking about hardware to talking about software...}
+Secondly, though AMD markets the processor as having 16 cores, it actually has only 8 floating-point units.
+As such, some applications that are floating-point intensive run with only 8 processes per node.
+This actually leaves 8 cores available so long as they primarily execute integer instructions.
 
-In most HPC applications, filesystem writes typically follow a very 'bursty' pattern.
-That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output when the application writes its checkpoint or output files.
-At ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it.
-This means that the performance that a user sees at any given time will likely vary dramatically.
-
-Having a sufficiently large write-back cache could improve applications' write performance in two ways.
-First, if the cache is large enough to hold the entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.
-It can write to the cache and then resume its computations while the cache drains to the filesystem.
-Secondly, write performance should be more predictable.
-From the application's point of view, write speed would only limited by how fast data can be copied into the cache, and that is not effected by outside influences.
-
- The GPU memory and integer cores can be used to implement such a cache without taking resources away from the main application.
-
-Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory.
-Its bandwidth is too low and its latency is too high.
-However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast.
-Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load.
-This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.
+ The unused GPU memory and integer cores can be used to implement a write-back cache without taking resources away from the main application.
+This paper demonstrates a technique for providing such a cache to applications.
+Runs were made on Titan using a synthetic benchmark to demonstrate performance both without such a cache and with a cache that utilizes ram on the GPU card.
+The results show definite performance benefits.
 
@@
 Performance benefits and any potential trade-offs and caveats are discussed.
+
+ Since the GPU memory is accessed via the PCIe bus, it is generally not practical to use it as part of a process's main memory.
+Its bandwidth is too low and its latency is too high.
+However, when compared to filesystem performance - even a very fast parallel filesystem - accessing this memory is still very fast.
+Tests on Titan show that writing to GPU memory is at least eight times faster than writing to the filesystem, and considerably better than that if the filesystem is under heavy load.
+This suggests that using this memory as a large write-back cache for file I/O might help application runtimes significantly.
+
 \end{comment}

### 2015-04-01 cug_paper/02_design.tex
Paper: Adding figures
Added great big figure showing the communications steps

@@
 
+\todo[inline]{Somewhere in here, we're going to want to insert the graph of host-to-gpu bandwidth vs.
+transfer size.
+It justifies chosing 16MB bocks and we're going to want to reference it from the analysis section, too.}
+
 \subsection{Overview}
@@
 This idea was tested using a synthetic benchmark application written by the authors.
-The benchmark consists of two executables: the main executable which simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
+The benchmark consists of two executables: the main executable which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
 The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.
@@
 
-With the daemon \todo{Should 'daemon' be plural?} started, the application begins its main execution.  In this case, since it's just a synthetic 
+With the daemon process running on each node, the application begins its main execution.  In this case, since it's just a synthetic 
 I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.
 For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.
+
+\begin{figure*}[!t]
+\centerline{
+  \subfloat[Application sends write request message to daemon]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide09}%
+    \label{fig:comm_a}}
+  \hfil
+  \subfloat[Daemon allocates GPU memory and converts pointer into IPC handle]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide10}%
+    \label{fig:comm_b}}}
+\centerline{
+  \subfloat[Daemon sends reply message with IPC handle]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide11}%
+    \label{fig:comm_c}}
+  \hfil
+  \subfloat[Application writes data to GPU memory]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide12}%
+    \label{fig:comm_d}}}    
+\centerline{
+  \subfloat[Application sends write complete message to daemon]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide13}%
+    \label{fig:comm_e}}
+  \hfil
+  \subfloat[Daemon copies data from GPU memory to system memory]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide14}%
+    \label{fig:comm_f}}} 
+\centerline{
+  \subfloat[Daemon writes data to filesystem]{
+    \includegraphics[width=2.5in]{figures/Data_Movement/Slide15}%
+    \label{fig:comm_g}}} 
+\caption{Communications steps}
+\label{fig_sim}
+\end{figure*}
 

### 2015-04-03 cug_paper/04_futurework.tex
Paper: Future Work section now at draft status

--- 2015-03-20 cug_paper/04_futurework.tex
@@
 
-Discuss work yet to be done:  add support for caching in system memory,  make 'production ready' (ie: easier for applications to use)
+The authors work in a group that is not dedicated to pure research; part of the group's mandate is to deploy useful software in the production environment.
+As such, the most important piece of future work is to convert this demonstration code into a package that can actually be used by other applications running on Titan.
+In the introduction, the authors noted that one of the reasons some applications haven't been converted to use the GPU for calculations is that the applications' developers haven't had time.
+With that in mind, it's impractical to expect those same developers to make major changes in order to use this code.
+
+The authors are considering a number of methods of packaging this code to make it easy for other developers to integrate into their own applications.
+One possibility is to intercept certain POSIX function calls, such as \texttt{open()} and \texttt{write()}, among others.
+While somewhat tricky to implement, this technique has been used with good success by the Mercury Posix project.\cite{mercury_posix_demo}  A second possibility is to modify one or more I/O libraries that are popular with applications, such as the NetCDF library, to make use of this code.
+This has the advantage of requiring zero code changes by the application developers.
+Properly maintaining the modified I/O libraries could be a real problem, though.
+A final possibility is to integrate this code into the Functional Partitioning project that the authors and their colleagues are working on.\cite{Li:2010:FPO:1884643.1884686}  Obviously, the authors could choose more than one of these alternatives.
+
+Besides deploying this code in a production environment, there are a few features the authors would like to add.
+The simplest is to add a user-controlled option to limit the amount of GPU memory the daemon will try to allocate.
+As currently written, the daemon will allocate all the memory it can.
+That could cause problems if the application wants to use the GPU for calculations.
+By allowing the user to set a hard limit on the amount of memory the daemon will use, it becomes possible for this software to coexist with applications that use the GPU for calculations.
+
+Another feature the authors would like to add is the ability to use regular system RAM in addition to the GPU RAM.
+It's well known that some applications don't use all the RAM on the compute node.
+It's debatable whether letting the daemon use the memory is better than simply letting the operating system use it as page cache.
+However, both the Linux kernel page cache and the client-side cache have fairly low limits on the amount of dirty pages they will allow.
+Some very basic initial testing implies that allowing the daemon to allocate memory may be more useful than letting the OS use the same memory for page cache, but further testing is required to be certain.
+
+Lastly, the next big supercomputer at ORNL has already been announced.
+While some details have yet to be finalized, one thing that has been decided is that each node will have approximately 800GB of nonvolatile memory (NVM).\cite{summit_page}
+Exactly how applications will use this NVM memory is the subject of much research and debate.
+With only minor changes, this software could write to the NVM instead of (or in addition to) the GPU memory.
+Assuming the ease-of-use goals mentioned above are met, the authors feel this would allow application developers to quickly modify their applications to use the NVM as a large burst-buffer, even as those developers look for other, more effective ways to make use of the hardware.

### 2015-04-03 analysis/plot_utils.py
More changes to the analysis code
Added some more helper functions for plotting per-rank bandwidth

@@ -19,6 +19,11 @@ def timelines(y, xstart, xstop, color='b'):
 
 TABLE_NAME="Results"
 
+
+# These 2 plot functions aren't used - we never could get the 
+# FCFS allocator to work properly
+
+'''
 # Plot average per-rank bandwidth as a function of write size
 # Assumes:
 #  First n/2 ranks used FCFS allocator, second n/2 ranks used balanced allocator
@@ -63,12 +68,11 @@ def plot_per_node_bw( conn, num_write_threads):
                       x_label = 'Write size (MB)',
                       y_label = 'Bandwidth (MB/s)',
                       legend=('First Come/First Served', 'Balanced'))
-
-
+                      
 # Bar chart of per-node BW vs. write size.  Assumes all ranks were
 # writing the normal way (not using the GPU cache)
 def plot_per_node_bw_non_gpu( conn, num_write_threads):
-    # Queries that we'll execute (the plot will show 2 groupings)
+    # Queries that we'll execute
     statements = ["SELECT length, SUM(bw) from %s GROUP BY node, iteration ORDER BY length;" \
                   %TABLE_NAME ]
     # See above for comments about the GROUP BY & SUM() clauses
@@ -77,6 +81,91 @@ def plot_per_node_bw_non_gpu( conn, num_write_threads):
                       title = 'Average Per-Node Bandwidth vs. Write Size',
                       x_label = 'Write size (MB)',
                       y_label = 'Bandwidth (MB/s)')
+
+
+'''
+
+
+# Plot average per-rank bandwidth as a function of write size
+# Assumes the first 512 ranks used 1 write thread, the next 512 used 2 write
+# threads and the last 512 used 4 write threads
+def plot_per_rank_bw_8rpn( conn, err_bars = True):
+    
+    #c = conn.cursor()
+    #c.execute( "SELECT max(rank) from %s;"%TABLE_NAME)
+    #max_rank = c.fetchall()[0][0]
+    
+    # Queries that we'll execute (the plot will show 2 groupings)
+    statements = [
+        "SELECT length, bw from %s WHERE rank < 512 ORDER BY length;"%TABLE_NAME,
+        "SELECT length, bw from %s WHERE rank < 1024 and rank >= 512 ORDER BY length;"%TABLE_NAME,
+        "SELECT length, bw from %s WHERE rank >=1024 ORDER BY length;"%TABLE_NAME ]
+    
+    generic_bar_plot( conn, statements, 
[…]

### 2015-04-04 cug_paper/03_analysis.tex
Paper: Analysis section now at rough draft state

--- 2015-03-20 cug_paper/03_analysis.tex
@@
 \subsection{Experiment Setup}
-Discuss what we ran, the parameters we swept and why we chose those parameters
+\label{subsec:exp_setup}
+
+In order to evaluate this software, the authors ran two series of tests: one that ran the application with 8 ranks per node and another that ran with 16 ranks per node.
+Each series included tests of the daemon using 1, 2 \& 4 background write threads.
+Both series also tested the write performance just using standard \texttt{write()} calls in order to get some baseline data for comparison.
+
+For the first test series, the application was configured for 8 ranks per node.
+As mentioned in the introduction, 'real' applications will often run on Titan using only 8 ranks per node because the AMD processor in the compute nodes only has 8 floating point units.
+From the authors' perspective, this has the advantage of leaving 8 integer-only cores to run the daemon in multi-threaded mode.
+For this test series, aprun was configured to pin the application ranks to the even numbered cores and the daemon was configured to pin its threads to odd numbered cores.
+Furthermore, the daemon thread that processed the CCI messages was run in a mode where it continuously polled for new messages.
+This  provided the lowest possible latency for message handling, but at the cost of effectively spinlocking one core.
+
+For the second test series, the application was configured for 16 ranks per node.
+The daemon was again run with 1, 2 \& 4 write threads.
+For this series, the daemon thread that handled the CCI messages was run in a mode where it would block waiting on a message.
+This added some latency to the message processing, but left that core free to perform useful work when there were no messages to process.
+Also for this series, no core pinning was used on the daemon.\footnote{In practical terms, a user would probably not want to use multiple write threads on the daemon if his/her applications was running 16 ranks/node since it would oversubscribe the cores.
+The authors tested the daemon with 2 and 4 threads partially out of curiosity and also to keep the two test series as similar to each other as possible.}
+
+For both test series, the daemon was configured to write each rank's data to a separate file and the test measured the application's perceived write performance as the write size increased.
+Note the word 'perceived'.
+What's being measured is how long it takes for each rank of the application to copy its data into the GPU memory.
+The GPU cards in each of Titan's compute nodes have 6GB of RAM, though in practice only a little over 5GB is actually available to the user.
+This means that for the first test series, with 8 ranks per node, write sizes of up to 512 MB were small enough for all ranks' writes to fit into the GPU memory.
+For the second test series, using 16 ranks per node, write sizes up to 256MB would fit.
+For both test series, once the write size exceeded the available GPU memory, the ranks would have to wait until the daemon was able to drain data out of the GPU memory and into the filesystem.
 
 \subsection{Results}
-Discuss the results
+\label{subsec:results}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/figure_1.eps}
+\caption{Average per-rank write bandwidth using standard \texttt{write()} calls.  8 ranks per node. No GPU memory cache.} 
+\label{fig:results_base_8}
+\end{figure}
+
+Figure \ref{fig:results_base_8} shows the results of a baseline test.
+The colored bars show the average values and the error bars display the minimum and maximum values.
+In this test, the application wrote to the filesystem with standard \texttt{write()} calls and the GPU memory was not used at all.\footnote{The daemon process was not even started for this test.}  The results provide some baseline numbers that can be used for comparison with the tests using GPU memory.
+
+Note the sharp drop in performance between 256MB and 512MB.
+On Titan, the Lustre client is configured to allow a maximum of 64MB per OST and to default to using 4 OST's per file.
+Given the number of OST's available, it's statistically likely that no two output files in this test used the same OST's.
+In short, write sizes of 256MB or less were cached in system memory using the existing Lustre client cache and the performance of the 512MB, 1GB and 2GB sizes is dominated by the performance of the filesystem.
+In order for this work to be useful, it must obviously improve on that.
+
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/figure_2.eps}
+\caption{Average per-rank write bandwidth.  8 ranks per node.} 
+\label{fig:results_8_nobars}
+\end{figure}
+
+Figure \ref{fig:results_8_nobars} shows the remaining results of the first test series.
+This series was run with 8 ranks per node and with 1, 2 \& 4 write threads in the daemon.
+The graph shows a number of interesting features.
+The first and most obvious, is that multiple write threads decrease performance.
+\todo[inline]{Should we just forget about multiple write threads?  I could re-do the plots easily enough to only include single thread data...}
+
+Concentrating on the single thread performance, it's clear that the application benefits from using the GPU memory out to the 512MB write size.
+This makes sense since eight ranks each writing 512MB is a total of 4GB and that will fit into the GPU memory.
+Even at 1GB, the application sees somewhat improved performance because there is enough GPU memory to hold significant fraction of the data to be written.
+It's not until the 2GB write size that the write performance is dominated by the filesystem's throughput.
+
+%\begin{figure}
+%\includegraphics[width=\linewidth]{figures/figure_3.eps}
+%\caption{Minimum \& maximum per-rank write bandwidth.  8 ranks per node.} 
+%\label{fig:results_8_bars}
+%\end{figure}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/figure_6.eps}
+\caption{Average per-rank write bandwidth using standard \texttt{write()} calls.  16 ranks per node. No GPU memory cache.} 
+\label{fig:results_base_16}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/figure_4.eps}
+\caption{Average per-rank write bandwidth.  16 ranks per node.} 
+\label{fig:results_16_nobars}
+\end{figure}
+
+%\begin{figure}
+%\includegraphics[width=\linewidth]{figures/figure_5.eps}
+%\caption{Minimum \& maximum per-rank write bandwidth.  16 ranks per node.} 
+%\label{fig:results_16_bars}
+%\end{figure}
+
+Figures \ref{fig:results_base_16} and \ref{fig:results_16_nobars} show the results for the second series of tests.
+As noted above, this series used 16 ranks per node, plus the daemon's threads.
+This meant, of course, that the cores were oversubscribed.
+Notice that Figure \ref{fig:results_base_16} as the same basic shape as Figure \ref{fig:results_base_8}.
+The only significant difference is that the reported speeds shown in Figure \ref{fig:results_base_16} are approximately half those shown in Figure \ref{fig:results_base_8}.
+This is expected, since there are twice as many ranks writing.
+
+Figure \ref{fig:results_16_nobars} shows that again, the per-rank bandwidth is about half that shown in Figure \ref{fig:results_8_nobars} because there are twice as many ranks.
+It's also clear that the application ranks see good write performance up to the 256MB write size.
+\todo[inline]{Need a little more analysis for the 16 rank-per-node tests.}
+
+Figures \ref{fig:results_base_8} and \ref{fig:results_base_16} show that Lustre client cache can provide good write performance up to 256MB write sizes.
+However, that presumes that there is, in fact, sufficient system memory available on the node \emph{and} that the individual ranks are writing to different OST's.
+If a 'real' application only wrote to a single file, the Lustre cache would not provide nearly as large a benefit unless the user took additional steps to increase the stripe count of the file.
+It's worth noting that if an application used the techniques demonstrated in this paper, it would see improved write performance regardless of how many OST's were used and how much system ram was available for cache.
+
+Looking at figure \ref{fig:results_16_nobars}, note the drop in performance between 16MB and 32MB and the slight downward trend from 32MB to 256MB.
+This pattern also appears in Figure \ref{fig:results_8_nobars}, but is much more obvious in Figure\ref{fig:results_16_nobars}.
+The authors hypothesize that this is related to the response time of the daemon.
+As mentioned earlier, the daemon allocates GPU memory in blocks of up to 16MB.
+Thus, if one of the ranks wants to write more than 16GB, it will need to go through more than one request/response message cycle.
+However, the daemon only has a single thread to handle these messages.
+When multiple ranks are sending message, they will necessarily have to wait on a response from the daemon and this slow their overall performance.
+This is more noticeable in Figure\ref{fig:results_16_nobars} because there are more ranks making requests.
+In short, it appears that improving the response time of the message handling thread would be beneficial.

### 2015-04-06 cug_paper/01_introduction.tex
Paper: Section 2 is now at rough draft stage
Fix single-quotes
To single quote foo, use `foo' instead of 'foo'

In some of these cases, we might want to drop single-quotes and use
\emph{}, \textbf{}, or \texttt{}

@@
 
-In most HPC applications, filesystem writes typically follow a very 'bursty' pattern.
+In most HPC applications, filesystem writes typically follow a very `bursty' pattern.
 That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.
@@
 Having a sufficiently large write-back cache would alleviate both of these problems.
-Assuming the cache is large enough, then the 'bursty' nature of the output pattern can be smoothed out since there will be time for the traffic to drain from the cache before the next burst arrives.
+Assuming the cache is large enough, then the `bursty' nature of the output pattern can be smoothed out since there will be time for the traffic to drain from the cache before the next burst arrives.
 Equally importantly, write performance should be more predictable.

### 2015-04-06 cug_paper/02_design.tex
Fix single-quotes
To single quote foo, use `foo' instead of 'foo'

In some of these cases, we might want to drop single-quotes and use
\emph{}, \textbf{}, or \texttt{}

@@
 
-Note that the data has not made it out to the filesystem yet; it's just sitting in GPU memory.  Upon receiving the 'write done' message from the application, the daemon now knows it's free to
+Note that the data has not made it out to the filesystem yet; it's just sitting in GPU memory.  Upon receiving the `write done' message from the application, the daemon now knows it's free to
 
@@
 
-From the application's point of view, the write has completed once the 'write done' message has been successfully sent.
+From the application's point of view, the write has completed once the `write done' message has been successfully sent.
 
-Once the applications sends the 'write done' message
+Once the applications sends the `write done' message
 
@@
 Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.
-This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called 'gpudefault'.
+This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.
 The naming scheme comes from the CUDA documentation.
@@
 Titan is configured differently, however.
-Thus the need to explicitly tell qsub to switch to the 'default' mode.}
+Thus the need to explicitly tell qsub to switch to the `default' mode.}
 

### 2015-04-06 cug_paper/02_design.tex

--- 2015-04-01 cug_paper/02_design.tex
@@
 \label{sec:design}
-
-\todo[inline]{Discuss how the software works.
-Things like using CCI for small messages, bulk transfers happening by way of cudaMemcpy().
-Mention Cray-specific stuff such as setting 'feature=gpudefault' in the qsub script.
-}
-
-\todo[inline]{Somewhere in here, we're going to want to insert the graph of host-to-gpu bandwidth vs.
-transfer size.
-It justifies chosing 16MB bocks and we're going to want to reference it from the analysis section, too.}
 
@@
 This idea was tested using a synthetic benchmark application written by the authors.
-The benchmark consists of two executables: the main executable which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
+The benchmark consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
 The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.
-Note that the daemon does \emph{not} copy the data into the GPU memory.
+Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application.
 There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.
@@
 For these tests, the sleep time was deliberately calculated to allow enough time for the data in GPU memory to drain.
+\subsection{Communication Between the Application and Daemon}
+\label{subsec:comm}
 
+Communication between the main application and the daemon happens over two different channels.
+Short messages for coordination between the application and the daemon are handled using the Common Communication Interface (CCI).\cite{atchley11:cci}  Bulk data is copied to and from GPU memory using \texttt{cudaMemcpy()}.
+
+\todo[inline]{We need to figure out how to force figure \ref{fig:communications} to display as page 2 or 3.  Right now, latex is putting it on the last page.}
 \begin{figure*}[!t]
@@
 \caption{Communications steps}
-\label{fig_sim}
+\label{fig:communications}
 \end{figure*}
 
-When it's time to write, the application sends a short message to the daemon.
+Writing data from the application is a seven step process.
+
+When it's time to write, the application sends a short message to the daemon using CCI.
 This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.
-For production code, something more flexible will obviously be needed.}
-\todo[inline]{We need a diagram showing the communication between the processes.}  The daemon will attempt to allocate memory on the GPU.
-The daemon will then reply to the application with the size of memory that was actually allocated and a memory handle that the application can use to access the GPU memory.
+For production code, something more flexible will obviously be needed.}(Figure \ref{fig:comm_a})
+
+When the daemon receives this message, it will attempt to allocate memory on the GPU.
+Assuming the allocation succeeds, the daemon will use \texttt{cudaIpcGetMemhandle()} to create a memory handle for the allocation.
+(Figure \ref{fig:comm_b})
+
+The daemon will then reply to the application with the size of memory that was actually allocated and the memory handle that the application can use to access the GPU memory.
 Note that the size value may be less than what the application asked for.
+(Figure \ref{fig:comm_c})
 
-Upon receiving the reply, the application opens the CUDA memory handle and copies the specified amount of data up to the GPU memory.
-When the copy has completed, the application closes the memory handle and sends a message back to the daemon say that this particular write has completed.
-If the application has more data to write, it can send another write request, or else it can continue with its calculations.
+Upon receiving the reply, the application opens the memory handle and converts it back to a standard pointer.
+The application then uses \texttt{cudaMemcpy()} to copy the specified amount of data up to the GPU memory.
+(Figure \ref{fig:comm_d})
 
-Note that the data has not made it out to the filesystem yet; it's just sitting in GPU memory.  Upon receiving the 'write done' message from the application, the daemon now knows it's free to
+When the copy has completed, the application closes the memory handle and sends a message back to the daemon saying that this particular write has completed.
+(Figure \ref{fig:comm_e})  If the application has more data to write, it can repeat the previous steps, or else it can continue with its calculations.
+
+Note that at this point the data has not made it out to the filesystem yet; it's just sitting in GPU memory.
+
+The daemon maintains a pair of lists.
+One list is the blocks of GPU memory that have been assigned to the ranks of the main application, and which the ranks are presumably copying data into.
+The second list is for all the blocks of memory for which the 'write complete' message (step e) has been received.
+When the daemon thread that handles CCI messages receives a 'write done' message for a block, it moves that block from the first list to the second list.
+It then notifies the background write thread(s) that the block is now ready to be written to the filesystem.
 
 The daemon maintains one or more background threads that are responsible for copying data out of GPU memory into normal system memory and then writing that data to the filesystem.
+These threads allow the writes to the filesystem to happen asynchronously while the main thread handles all the CCI messages.
+Since it's impossible to write directly from GPU memory to the filesystem, each thread will copy a block of data from GPU memory into a buffer in main memory.
+(Figure \ref{fig:comm_f})  It will then write the data to the output file at the location specified in the original request message.
+(Figure \ref{fig:comm_g})  Once data has been written to the filesystem, the write thread free's the GPU memory so that it can be used in another request.
 
-The background write thread(s) operates asynchronously.
+\begin{figure}
+\includegraphics[width=\linewidth]{figures/Host2Device_BW.pdf}
+\caption{Bandwidth between system memory and GPU memory as a function of write size\cite{ac_guide}} 
+\label{fig:transfer_bw}
+\end{figure}
 
-Each thread copies data out of GPU memory and into a buffer in system memory.
-From there, it is written out to the filesystem.
-The system memory buffer is deliberation kept fairly small - 16MB - in order to conserve resources.
-Testing done on Titan shows that 16MB block sizes are large enough for cudaMemcpy() to give good performance.\cite{ac_guide}  Larger buffers would simply use up memory without improving performance noticeably.
+Note that each write thread in the daemon has its own buffer in system memory.
+Each of these buffers is deliberately kept fairly small - 16MB - in order to conserve resources.
+Testing done on Titan shows that 16MB block sizes are large enough for \texttt{cudaMemcpy()} to give good performance.\cite{ac_guide}  Larger buffers would simply use up memory without improving performance noticeably.
+See Figure \ref{fig:transfer_bw}.
 
-\todo[inline]{Do we need to get into the details of cuda IPC functions? cudaIpcGetMemHandle, cudaIpcOpenMemHandle, cudaIpcCloseMemHandle}
-
-From the application's point of view, the write has completed once the 'write done' message has been successfully sent.
-
-Once the applications sends the 'write done' message
-
-\todo[inline]{Need to mention CCI - short messages, low latency.  Citation reference is "atchley11:cci"}
-
-The GPU attempts to perform some load-balancing by restricting the amount of memory allocated for any single request.
-Currently this limit is 16MB.
+Also note that the daemon limits the amount of memory allocated to any single request to 16MB.
+This was done to provide some limited load-balancing.
 If an application wants to write 512MB, it will therefore have to make 32 separate requests.
-If other ranks on the node are trying to write at the same time, their requests will all be interleaved and hopefully rank will be starved.
-
-It's debatable whether this sort of load balancing is necessary or even desirable.
-Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).
-However, that same testing showed that one of the CUDA functions - cudaIpcOpenMemHandle() - became unreliable once GPU memory allocations exceeded 512MB.
-Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.
+If other ranks on the node are trying to write at the same time, their requests will all be interleaved and hopefully no rank will be starved out.
 
@@
 
-\todo[inline]{Need an introduction paragraph for this subsection.}
+There are a few peculiarities of the Cray environment that need to be taken in to account in order for this software to work properly.
+The first peculiarity has to do with limiting access to the GPU and associated hardware.
 The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.
@@
 Titan is configured differently, however.
-Thus the need to explicitly tell qsub to switch to the 'default' mode.}
+Thus the need to explicitly tell qsub to switch the GPU's to the 'default' mode.}
 
-\todo[inline] {talk about actually starting up the daemon - fork(), open(excl), exec() }
+The second peculiarity has to do with actually starting the daemon process.
+Normally, executables are started on the compute nodes by aprun.
+However, aprun is designed to start a single executable, not to start two different executables.
+Even in MPMD mode, it won't start two different executables on the same node.
+The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.
+Since MPI purposely hides the compute topology from the application, deciding exactly which ranks call fork() takes a little more work.
+Specifically, all ranks will attempt to open a file using the \texttt{O\_CREAT} and
+\texttt{O\_EXCL} flags.
+The name of the file is taken from the compute node's host name.
+This combination of flags and filename ensures that exactly one rank on each node successfully opens the file.
+The ranks that succeed will then start the daemon processes.
+
+%
+% Everything below is commented out.  It's text I've toyed with using, but decided not to for whatever reason
+%
+\begin{comment}
+
+It's debatable whether this sort of load balancing is necessary or even desirable.
+Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).
+However, that same testing showed that one of the CUDA functions - \texttt{cudaIpcOpenMemHandle()} - became unreliable once GPU memory allocations exceeded 512MB.
+Until the nature of that problem could be determined, it was decided to limit allocations to 16MB.
+
+CUDA has several functions for facilitating interprocess communications.
+Among other things, one can convert a device pointer to a process-independant memory handle.
+Another process can take that memory handle and convert it back to a device pointer.
+
+\end{comment}

### 2015-04-06 cug_paper/02_design.tex
Fix merge conflicts

@@
 One list is the blocks of GPU memory that have been assigned to the ranks of the main application, and which the ranks are presumably copying data into.
-The second list is for all the blocks of memory for which the 'write complete' message (step e) has been received.
-When the daemon thread that handles CCI messages receives a 'write done' message for a block, it moves that block from the first list to the second list.
+The second list is for all the blocks of memory for which the `write complete' message (step e) has been received.
+When the daemon thread that handles CCI messages receives a `write done' message for a block, it moves that block from the first list to the second list.
 It then notifies the background write thread(s) that the block is now ready to be written to the filesystem.
@@
 Titan is configured differently, however.
-Thus the need to explicitly tell qsub to switch the GPU's to the 'default' mode.}
+Thus the need to explicitly tell qsub to switch the GPU's to the `default' mode.}
 

### 2015-04-06 cug_paper/01_introduction.tex
Scott's intro edits

@@
 
-In most HPC applications, filesystem writes typically follow a very `bursty' pattern.
+Most HPC applications' I/O exhibit a very `bursty' pattern.
 That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.
-Also, at ORNL, the filesystem is a shared resource and other users will also be putting a varying load on it.
-This means that the performance that a user sees at any given time will likely vary dramatically.
+At ORNL, the filesystem is a center-wide resource shared with other systems including a Cray XC-30, analysis clusters, and WAN-connected data mover nodes.
+Each of these competing systems place a varying load on the filesystem such that Titan's users' write performance will vary dramatically within a job and between jobs.
 
-Having a sufficiently large node-local write-back cache could improve applications' performance in two ways.
+Having a sufficiently large, node-local, write-back cache could improve an application's performance in two ways.
 First, if the cache is large enough to hold an entire write and the individual writes are far enough apart, then the application never has to wait on the filesystem.
@@
 Titan, the Cray XK7 at Oak Ridge National Laboratory, has 18,688 compute nodes.
-Each node consists of a 16-core AMD CPU, an NVIDIA GPU and 32GB ram.
-In addition, there is another 6GB of ram on each GPU card.
-Not surprisingly, there are applications that do not use all the resources on a compute node.
+Each node consists of a 16-core AMD Interlagos CPU, an NVIDIA GPU and 32GB DRAM.
+In addition, there is another 6GB of GDDR DRAM on each GPU card.
+Not surprisingly, some applications do not use all the resources on a compute node.
 
@@
 There may be good reasons for this.
-Utilizing the GPU's requires changes to the application and it's possible the developers haven't had the time to do that.
-It's also possible that the type of computations the application performs simply don't lend themselves to GPU-based acceleration.
-Regardless of why, it's clear that not all applications are benefiting from the GPU.
+Utilizing the GPU requires changes to the application and it is possible the developers have not had the time or funding to do that.
+It is also possible that the type of computations the application performs simply do not lend themselves to GPU-based acceleration (i.e.
+not enough exposed parallelism).
+Regardless of why, it is clear that not all applications are benefiting from the GPU.
 
-Secondly, though AMD markets the processor as having 16 cores, it actually has only 8 floating-point units.
-As such, some applications that are floating-point intensive run with only 8 processes per node.
-This actually leaves 8 cores available so long as they primarily execute integer instructions.
+Secondly, the Interlagos has eight Bulldozer units, each of which has two full integer units and a shared floating-point unit.
+Some users with floating-point intensive applications run jobs using only eight processes per node (one process per Bulldozer) to avoid oversubscribing the floating-point units.
+This actually leaves eight integer cores available so long as they primarily execute integer instructions.
 
@@
 This paper demonstrates a technique for providing such a cache to applications.
-Runs were made on Titan using a synthetic benchmark to demonstrate performance both without such a cache and with a cache that utilizes ram on the GPU card.
+Runs were made on Titan using a synthetic benchmark to demonstrate performance both without such a cache and with a cache that utilizes DRAM on the GPU card.
 The results show definite performance benefits.
 
+%
+% Everything below is commented out.  It's text I've toyed with using, but decided not to for whatever reason
+%
 \begin{comment}

### 2015-04-06 cug_paper/02_design.tex
Fix figure 1 placement
Remove comment about figure 1 placement
Paper:  Add a footnote discussing the difference between pinned mem and pageable mem

@@
 
-\todo[inline]{We need to figure out how to force figure \ref{fig:communications} to display as page 2 or 3.  Right now, latex is putting it on the last page.}
-\begin{figure*}[!t]
+\begin{figure*}[hp]
 \centerline{
@@
 Testing done on Titan shows that 16MB block sizes are large enough for \texttt{cudaMemcpy()} to give good performance.\cite{ac_guide} Larger buffers would simply use up memory without improving performance noticeably.
-See Figure \ref{fig:transfer_bw}.
+(See Figure \ref{fig:transfer_bw}.\footnote{Figure \ref{fig:transfer_bw} differentiates between 'pinned' memory and 'pageable' memory.
+Pageable memory is allocated using regular \texttt{malloc()} or \texttt{new} calls.
+Pinned memory is allocated using \texttt{cudaMallocHost()} or \texttt{cudaHostAlloc()}.\cite{cuda_pinned_memory}  This work uses pinned memory exclusively.})
 

### 2015-04-06 cug_paper/02_design.tex
Replace fig 1 with two figures

@@
 \label{subsec:overview}
-This idea was tested using a synthetic benchmark application written by the authors.
-The benchmark consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
+To test this idea, we created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
 The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.
 Note that the daemon does \emph{not} copy the data into the GPU memory; that is handled by the individual ranks of the main application.
-There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.
-\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.
+There needs to be exactly one daemon process on each compute node regardless of how many MPI ranks are running on each node.\footnote{Due to the limitations of aprun, starting this daemon is somewhat involved.
 See section \ref{subsec:cray-specific}.}
 
-With the daemon process running on each node, the application begins its main execution.  In this case, since it's just a synthetic 
+With the daemon process running on each node, the application begins its main execution.  In this case, since it is just a synthetic 
 I/O benchmark, it performs no actual computation; it simply writes a specified amount of data and then sleeps for a specified amount of time.
@@
 
-Communication between the main application and the daemon happens over two different channels.
-Short messages for coordination between the application and the daemon are handled using the Common Communication Interface (CCI).\cite{atchley11:cci}  Bulk data is copied to and from GPU memory using \texttt{cudaMemcpy()}.
+The main application and daemon coordinate with each other using short messages provided by the Common Communication Interface (CCI).\cite{atchley11:cci} Bulk data is copied to and from GPU memory using \texttt{cudaMemcpy()}.
 
-\begin{figure*}[hp]
-\centerline{
-  \subfloat[Application sends write request message to daemon]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide09}%
-    \label{fig:comm_a}}
-  \hfil
-  \subfloat[Daemon allocates GPU memory and converts pointer into IPC handle]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide10}%
-    \label{fig:comm_b}}}
-\centerline{
-  \subfloat[Daemon sends reply message with IPC handle]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide11}%
-    \label{fig:comm_c}}
-  \hfil
-  \subfloat[Application writes data to GPU memory]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide12}%
-    \label{fig:comm_d}}}    
-\centerline{
-  \subfloat[Application sends write complete message to daemon]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide13}%
-    \label{fig:comm_e}}
-  \hfil
-  \subfloat[Daemon copies data from GPU memory to system memory]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide14}%
-    \label{fig:comm_f}}} 
-\centerline{
-  \subfloat[Daemon writes data to filesystem]{
-    \includegraphics[width=2.5in]{figures/Data_Movement/Slide15}%
-    \label{fig:comm_g}}} 
-\caption{Communications steps}
-\label{fig:communications}
-\end{figure*}
+\begin{figure}
+\includegraphics[trim = .75in .75in .75in .75in, clip, width=0.5\textwidth]{figures/Data_Movement/Slide20}%
+\caption{Buffering data to GPU memory}
+\label{fig:buffer_write}
+\end{figure}
+
+\begin{figure}
+\includegraphics[trim = .75in .75in .75in .75in, clip, width=0.5\textwidth]{figures/Data_Movement/Slide21}%
+\caption{Draining data from GPU memory}
+\label{fig:buffer_drain}
+\end{figure}
+
+%\begin{figure*}[hp]
+%\centerline{
+%  \subfloat[Application sends write request message to daemon]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide09}%
+%    \label{fig:comm_a}}
+%  \hfil
+%  \subfloat[Daemon allocates GPU memory and converts pointer into IPC handle]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide10}%
+%    \label{fig:comm_b}}}
+%\centerline{
+%  \subfloat[Daemon sends reply message with IPC handle]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide11}%
+%    \label{fig:comm_c}}
+%  \hfil
+%  \subfloat[Application writes data to GPU memory]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide12}%
+%    \label{fig:comm_d}}}    
+%\centerline{
+%  \subfloat[Application sends write complete message to daemon]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide13}%
+%    \label{fig:comm_e}}
+%  \hfil
+%  \subfloat[Daemon copies data from GPU memory to system memory]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide14}%
+%    \label{fig:comm_f}}} 
+%\centerline{
+%  \subfloat[Daemon writes data to filesystem]{
+%    \includegraphics[width=2.5in]{figures/Data_Movement/Slide15}%
+%    \label{fig:comm_g}}} 
+%\caption{Communications steps}
+%\label{fig:communications}
+%\end{figure*}
 
 Writing data from the application is a seven step process.
-
-When it's time to write, the application sends a short message to the daemon using CCI.
+When it is time to write, the application sends a short message to the daemon using CCI.
 This message is a request to write data, and contains the amount of data and the starting offset into the output file.\footnote{For this benchmark, the daemon names output files based on the MPI rank of the requestor.
-For production code, something more flexible will obviously be needed.}(Figure \ref{fig:comm_a})
+For production code, something more flexible will obviously be needed.} (Figure \ref{fig:buffer_write}a)
 
@@
 Assuming the allocation succeeds, the daemon will use \texttt{cudaIpcGetMemhandle()} to create a memory handle for the allocation.
-(Figure \ref{fig:comm_b})
+(Figure \ref{fig:buffer_write}b)
 
@@
 Note that the size value may be less than what the application asked for.
-(Figure \ref{fig:comm_c})
+(Figure \ref{fig:buffer_write}c)
 
@@
 The application then uses \texttt{cudaMemcpy()} to copy the specified amount of data up to the GPU memory.
-(Figure \ref{fig:comm_d})
+(Figure \ref{fig:buffer_write}d)
 
 When the copy has completed, the application closes the memory handle and sends a message back to the daemon saying that this particular write has completed.
-(Figure \ref{fig:comm_e})  If the application has more data to write, it can repeat the previous steps, or else it can continue with its calculations.
+(Figure \ref{fig:buffer_write}e)  If the application has more data to write, it can repeat the previous steps, or else it can continue with its calculations.
 
-Note that at this point the data has not made it out to the filesystem yet; it's just sitting in GPU memory.
+Note that at this point the data has not made it out to the filesystem yet; it is just sitting in GPU memory.
 
@@
 These threads allow the writes to the filesystem to happen asynchronously while the main thread handles all the CCI messages.
-Since it's impossible to write directly from GPU memory to the filesystem, each thread will copy a block of data from GPU memory into a buffer in main memory.
-(Figure \ref{fig:comm_f})  It will then write the data to the output file at the location specified in the original request message.
-(Figure \ref{fig:comm_g})  Once data has been written to the filesystem, the write thread free's the GPU memory so that it can be used in another request.
+Since it is impossible to write directly from GPU memory to the filesystem, each thread will copy a block of data from GPU memory into a buffer in main memory.
+(Figure \ref{fig:buffer_drain}f)  It will then write the data to the output file at the location specified in the original request message.
+(Figure \ref{fig:buffer_drain}g)  Once data has been written to the filesystem, the write thread frees the GPU memory so that it can be used in another request.
 
@@
 However, aprun is designed to start a single executable, not to start two different executables.
-Even in MPMD mode, it won't start two different executables on the same node.
+Even in MPMD mode, it will not start two different executables on the same node.
 The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.
@@
 
-It's debatable whether this sort of load balancing is necessary or even desirable.
+It is debatable whether this sort of load balancing is necessary or even desirable.
 Indeed, preliminary testing seemed to indicate that it might be faster to let the daemon allocate  as much GPU memory as the client requested (assuming there was enough free).

### 2015-04-06 cug_paper/03_analysis.tex
Paper:  Added text for the conclusion.  Also added a little text to the analysis and moved the conclusion before the future work section

@@
 The first and most obvious, is that multiple write threads decrease performance.
-\todo[inline]{Should we just forget about multiple write threads?  I could re-do the plots easily enough to only include single thread data...}
+Exactly why this is so is unclear, but it appears that having multiple threads read from the GPU's memory interfered with the individual ranks' ability to write to the GPU's memory.
 
@@
 It's not until the 2GB write size that the write performance is dominated by the filesystem's throughput.
+
+Also obvious from the graph is the fact that small write sizes are not particularly efficient.
+For write sizes less than 4MB, copying data to the GPU memory is slower than writing to the Lustre cache.
+This is not surprising given the results shown in Figure \ref{fig:transfer_bw}.
 
@@
 Figure \ref{fig:results_16_nobars} shows that again, the per-rank bandwidth is about half that shown in Figure \ref{fig:results_8_nobars} because there are twice as many ranks.
-It's also clear that the application ranks see good write performance up to the 256MB write size.
-\todo[inline]{Need a little more analysis for the 16 rank-per-node tests.}
+It's also clear that the application ranks see good write performance up to the 256MB write size, which is the largest size that will entirely fit into the GPU's memory.
+It's also clear that, as in the first test series, writes must be at least 4MB in order to get reasonable performance copying the data into the GPU's memory.
 

### 2015-04-06 cug_paper/05_conclusion.tex
Paper:  Lots of work on the CUG paper.
Paper:  Added text for the conclusion.  Also added a little text to the analysis and moved the conclusion before the future work section

@@
 
-Discuss what we learned; summarize
+For the first test series (8 ranks/node), the results show clear performance improvement compared to writing directly to the filesystem.
+More data can be cached in the GPU's memory than in the Lustre client-side cache and the average write speed is, surprisingly, slightly higher.
+It is, however, necessary to use relatively large writes.
+As shown in Figure \ref{fig:transfer_bw}, the transfer bandwidth between the host and GPU is relatively poor for small transfers.
+Comparing Figure \ref{fig:results_base_8} and Figure \ref{fig:results_8_nobars},  it can be seen that write sizes must be at least 4MB before transfers to the GPU memory outperform writes to the Lustre cache.
+
+Note that these tests were run with file striping left at its default value of 4.
+In theory, writes to the filesystem could be improved by increasing the number stripes for each file, but that requires more system memory, which necessarily leaves less for the application.
+
+The results for the second test series (16 ranks/node) are less clear.
+As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into GPU RAM perform slightly better than just writing to disk.
+However, write sizes that exceed the capacity of the GPU RAM are actually a little slower than writing straight to the filesystem.
+
+There are several other complications to consider, though.
+First, Since the test application did no actual computation, there were plenty of cores available for the daemon to execute on.
+In a real HPC application, this wouldn't be the case and the application's performance would almost certainly be negatively affected by the context switches.
+On the other hand, an application that is running 16 ranks/node is unlikely to leave much system memory free for use by the Lustre client-side cache.
+So, caching writes in the GPU memory might offset the disruption caused by oversubscribing the cores.
+Whether or not this technique provides a net performance improvement in such a configuration will probably depend on the particular application.
+
+An interesting compromise might be to have the application run 15 ranks/node instead of 16.
+That will leave 1 core available for the daemon.
+Both threads could be pinned to that core, with the message processing thread given higher priority.
+It would also be possible to use the  core specialization option to aprun to schedule OS tasks on that core.
+Compared to an application running lots of MPI collectives, the daemon isn't particularly sensitive to short interrupts like that.
+This configuration hasn't been tested and the authors are unaware of any application that routinely runs with 15 ranks/node.
+
+In general, the results show that using the GPU memory for cache provides approximately the same performance as the Lustre client-side cache.
+The difference is that in order to benefit from the Lustre cache, the writes must be spread across a sufficient number of OST's and the memory must actually be available.
+Whether or not these conditions can be met depends on the particular application.

### 2015-04-06 cug_paper/02_design.tex
Scott's design edits

@@
 It then notifies the background write thread(s) that the block is now ready to be written to the filesystem.
+All daemon threads are pinned to unused integer cores.
 
@@
 Note that each write thread in the daemon has its own buffer in system memory.
-Each of these buffers is deliberately kept fairly small - 16MB - in order to conserve resources.
+Each of these buffers is deliberately kept fairly small, 16MB in this test, in order to conserve resources.
 Testing done on Titan shows that 16MB block sizes are large enough for \texttt{cudaMemcpy()} to give good performance.\cite{ac_guide} Larger buffers would simply use up memory without improving performance noticeably.
-(See Figure \ref{fig:transfer_bw}.\footnote{Figure \ref{fig:transfer_bw} differentiates between 'pinned' memory and 'pageable' memory.
+(See Figure \ref{fig:transfer_bw}.\footnote{Figure \ref{fig:transfer_bw} differentiates between `pinned' memory and `pageable' memory.
 Pageable memory is allocated using regular \texttt{malloc()} or \texttt{new} calls.
@@
 The first peculiarity has to do with limiting access to the GPU and associated hardware.
-The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process or to multiple threads from multiple processes.
+The CUDA runtime can be configured to limit access to the GPU to a single thread, to multiple threads from a single process, or to multiple threads from multiple processes.
 On Titan, if no special options are given to qsub, the CUDA runtime will be configured for a single thread.
 Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.
-This is done by passing the flag "-l feature=gpudefault" to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.
+This is done by passing the flag ``l feature=gpudefault'' to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.
 The naming scheme comes from the CUDA documentation.
@@
 The second peculiarity has to do with actually starting the daemon process.
-Normally, executables are started on the compute nodes by aprun.
-However, aprun is designed to start a single executable, not to start two different executables.
-Even in MPMD mode, it will not start two different executables on the same node.
+Normally, executables are started on the compute nodes by \emph{aprun}, Cray's replacement for mpirun.
+by default, aprun is designed to start a single executable in single-program-multiple-data (SPMD) mode, not to start two different executables.
+Aprun does support multiple-program-multiple-data (MPMD) mode, but it will not start two different executables on the same node (i.e.
+each program executes on non-overlapping subsets of nodes).
 The solution is to have one rank on each node call \texttt{fork()} and then \texttt{execve()}.

### 2015-04-06 cug_paper/03_analysis.tex
Paper: Grammar & punctuation editing for analysis & conclusion sections

@@
 Note the word 'perceived'.
-What's being measured is how long it takes for each rank of the application to copy its data into the GPU memory.
+What was actually measured was how long it took for each rank of the application to copy its data into the GPU's memory.
 The GPU cards in each of Titan's compute nodes have 6GB of RAM, though in practice only a little over 5GB is actually available to the user.
@@
 Figure \ref{fig:results_8_nobars} shows the remaining results of the first test series.
-This series was run with 8 ranks per node and with 1, 2 \& 4 write threads in the daemon.
+As stated earlier, this series was run with 8 ranks per node and with 1, 2 \& 4 write threads in the daemon.
 The graph shows a number of interesting features.
@@
 This meant, of course, that the cores were oversubscribed.
-Notice that Figure \ref{fig:results_base_16} as the same basic shape as Figure \ref{fig:results_base_8}.
+Notice that Figure \ref{fig:results_base_16} has the same basic shape as Figure \ref{fig:results_base_8}.
 The only significant difference is that the reported speeds shown in Figure \ref{fig:results_base_16} are approximately half those shown in Figure \ref{fig:results_base_8}.
@@
 
-Figures \ref{fig:results_base_8} and \ref{fig:results_base_16} show that Lustre client cache can provide good write performance up to 256MB write sizes.
-However, that presumes that there is, in fact, sufficient system memory available on the node \emph{and} that the individual ranks are writing to different OST's.
-If a 'real' application only wrote to a single file, the Lustre cache would not provide nearly as large a benefit unless the user took additional steps to increase the stripe count of the file.
-It's worth noting that if an application used the techniques demonstrated in this paper, it would see improved write performance regardless of how many OST's were used and how much system ram was available for cache.
-
 Looking at figure \ref{fig:results_16_nobars}, note the drop in performance between 16MB and 32MB and the slight downward trend from 32MB to 256MB.
-This pattern also appears in Figure \ref{fig:results_8_nobars}, but is much more obvious in Figure\ref{fig:results_16_nobars}.
+This pattern also appears in Figure \ref{fig:results_8_nobars}, but is much more obvious in Figure
+\ref{fig:results_16_nobars}.
 The authors hypothesize that this is related to the response time of the daemon.
@@
 However, the daemon only has a single thread to handle these messages.
-When multiple ranks are sending message, they will necessarily have to wait on a response from the daemon and this slow their overall performance.
-This is more noticeable in Figure\ref{fig:results_16_nobars} because there are more ranks making requests.
+When multiple ranks are sending message, they will necessarily have to wait while the daemon services any previous messages and this slows their overall performance.
+This is more noticeable in Figure
+\ref{fig:results_16_nobars} because there are more ranks making requests.
 In short, it appears that improving the response time of the message handling thread would be beneficial.

### 2015-04-06 cug_paper/05_conclusion.tex
Paper: Grammar & punctuation editing for analysis & conclusion sections

@@
 
-For the first test series (8 ranks/node), the results show clear performance improvement compared to writing directly to the filesystem.
+For the first test series (8 ranks/node), the results show a clear performance improvement compared to writing directly to the filesystem.
 More data can be cached in the GPU's memory than in the Lustre client-side cache and the average write speed is, surprisingly, slightly higher.
 It is, however, necessary to use relatively large writes.
-As shown in Figure \ref{fig:transfer_bw}, the transfer bandwidth between the host and GPU is relatively poor for small transfers.
 Comparing Figure \ref{fig:results_base_8} and Figure \ref{fig:results_8_nobars},  it can be seen that write sizes must be at least 4MB before transfers to the GPU memory outperform writes to the Lustre cache.
@@
 The results for the second test series (16 ranks/node) are less clear.
-As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into GPU RAM perform slightly better than just writing to disk.
-However, write sizes that exceed the capacity of the GPU RAM are actually a little slower than writing straight to the filesystem.
+As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into the GPU's memory perform slightly better than just writing to disk.
+However, write sizes that exceed the capacity of the GPU's memory are actually a little slower than writing straight to the filesystem.
 
@@
 On the other hand, an application that is running 16 ranks/node is unlikely to leave much system memory free for use by the Lustre client-side cache.
-So, caching writes in the GPU memory might offset the disruption caused by oversubscribing the cores.
+So, the improvement from caching writes in the GPU's memory might offset the disruption caused by oversubscribing the cores.
 Whether or not this technique provides a net performance improvement in such a configuration will probably depend on the particular application.
@@
 
-In general, the results show that using the GPU memory for cache provides approximately the same performance as the Lustre client-side cache.
+In general, the results show that using the GPU's memory to cache filesystem writes provides approximately the same performance as the Lustre client-side cache.
 The difference is that in order to benefit from the Lustre cache, the writes must be spread across a sufficient number of OST's and the memory must actually be available.
 Whether or not these conditions can be met depends on the particular application.
+
+%Figures \ref{fig:results_base_8} and \ref{fig:results_base_16} show that Lustre client cache can provide good write performance up to 256MB write sizes.
+However, that presumes that there is, in fact, sufficient system memory available on the node \emph{and} that data is being written to a sufficient number OST's.
+If a 'real' application only wrote to a single file, the Lustre cache would not provide nearly as large a benefit unless the user took additional steps to increase the stripe count of the file.
+It's worth noting that if an application used the techniques demonstrated in this paper, it would see improved write performance regardless of how many OST's were used and how much system ram was available for cache.

### 2015-04-06 cug_paper/04_futurework.tex

@@
 As such, the most important piece of future work is to convert this demonstration code into a package that can actually be used by other applications running on Titan.
-In the introduction, the authors noted that one of the reasons some applications haven't been converted to use the GPU for calculations is that the applications' developers haven't had time.
-With that in mind, it's impractical to expect those same developers to make major changes in order to use this code.
+In the introduction, the authors noted that one of the reasons some applications have not been converted to use the GPU for calculations is that the scientific applications' developers have not had time and/or funding.
+With that in mind, it is impractical to expect those same developers to make major changes in order to use this code.
 
-The authors are considering a number of methods of packaging this code to make it easy for other developers to integrate into their own applications.
+The authors are considering a number of methods of packaging this code to make it easy for application developers to integrate into their own applications.
 One possibility is to intercept certain POSIX function calls, such as \texttt{open()} and \texttt{write()}, among others.
@@
 Another feature the authors would like to add is the ability to use regular system RAM in addition to the GPU RAM.
-It's well known that some applications don't use all the RAM on the compute node.
-It's debatable whether letting the daemon use the memory is better than simply letting the operating system use it as page cache.
+It is well known that some applications do not use all the RAM on the compute node.
+It is debatable whether letting the daemon use the memory is better than simply letting the operating system use it as page cache.
 However, both the Linux kernel page cache and the client-side cache have fairly low limits on the amount of dirty pages they will allow.

### 2015-04-06 cug_paper/03_analysis.tex
Paper:  More edits & proofreading.  Last of the /todo boxes removed.

@@
 Note the word 'perceived'.
-What was actually measured was how long it took for each rank of the application to copy its data into the GPU's memory.
+What was actually measured was how long it took for each rank of the application to copy its data into GPU memory.
 The GPU cards in each of Titan's compute nodes have 6GB of RAM, though in practice only a little over 5GB is actually available to the user.
-This means that for the first test series, with 8 ranks per node, write sizes of up to 512 MB were small enough for all ranks' writes to fit into the GPU memory.
+This means that for the first test series, with 8 ranks per node, write sizes of up to 512 MB were small enough for all ranks' writes to fit into GPU memory.
 For the second test series, using 16 ranks per node, write sizes up to 256MB would fit.
-For both test series, once the write size exceeded the available GPU memory, the ranks would have to wait until the daemon was able to drain data out of the GPU memory and into the filesystem.
+For both test series, once the write size exceeded the available GPU memory, the ranks would have to wait until the daemon was able to drain data out of GPU memory and into the filesystem.
 
@@
 The first and most obvious, is that multiple write threads decrease performance.
-Exactly why this is so is unclear, but it appears that having multiple threads read from the GPU's memory interfered with the individual ranks' ability to write to the GPU's memory.
+Exactly why this is so is unclear, but it appears that having multiple threads read from GPU memory interfered with the individual ranks' ability to copy data into GPU memory.
 
-Concentrating on the single thread performance, it's clear that the application benefits from using the GPU memory out to the 512MB write size.
-This makes sense since eight ranks each writing 512MB is a total of 4GB and that will fit into the GPU memory.
+Concentrating on the single thread performance, it's clear that the application benefits from using GPU memory out to the 512MB write size.
+This makes sense since eight ranks each writing 512MB is a total of 4GB and that will fit into the available GPU memory.
 Even at 1GB, the application sees somewhat improved performance because there is enough GPU memory to hold significant fraction of the data to be written.
@@
 Also obvious from the graph is the fact that small write sizes are not particularly efficient.
-For write sizes less than 4MB, copying data to the GPU memory is slower than writing to the Lustre cache.
+For write sizes less than 4MB, copying data to GPU memory is slower than writing to the Lustre cache.
 This is not surprising given the results shown in Figure \ref{fig:transfer_bw}.
@@
 Figure \ref{fig:results_16_nobars} shows that again, the per-rank bandwidth is about half that shown in Figure \ref{fig:results_8_nobars} because there are twice as many ranks.
-It's also clear that the application ranks see good write performance up to the 256MB write size, which is the largest size that will entirely fit into the GPU's memory.
-It's also clear that, as in the first test series, writes must be at least 4MB in order to get reasonable performance copying the data into the GPU's memory.
+It's also clear that the application ranks see good write performance up to the 256MB write size, which is the largest size that will entirely fit into GPU memory.
+It's also clear that, as in the first test series, writes must be at least 4MB in order to get reasonable performance copying the data into GPU memory.
 

### 2015-04-06 cug_paper/04_futurework.tex
Paper:  More edits & proofreading.  Last of the /todo boxes removed.

@@
 It is well known that some applications do not use all the RAM on the compute node.
-It is debatable whether letting the daemon use the memory is better than simply letting the operating system use it as page cache.
-However, both the Linux kernel page cache and the client-side cache have fairly low limits on the amount of dirty pages they will allow.
-Some very basic initial testing implies that allowing the daemon to allocate memory may be more useful than letting the OS use the same memory for page cache, but further testing is required to be certain.
+It is debatable whether letting the daemon use the memory is better than simply letting the operating system use it as  Lustre client cache or OS page cacher.
+However, as was discussed in Section \ref{sec:analysis} and Section \ref{sec:conclusion}, the Lustre client-side cache has fairly low limits on the amount of dirty pages it will allow.
+(The Linux kernel page cache has similar low limits on dirty pages.)  Some very basic initial testing implies that allowing the daemon to allocate memory may be more useful than letting the OS use the same memory for page cache, but further testing is required to be certain.
 

### 2015-04-06 cug_paper/05_conclusion.tex
Paper:  More edits & proofreading.  Last of the /todo boxes removed.

@@
 For the first test series (8 ranks/node), the results show a clear performance improvement compared to writing directly to the filesystem.
-More data can be cached in the GPU's memory than in the Lustre client-side cache and the average write speed is, surprisingly, slightly higher.
+More data can be cached in GPU memory than in the Lustre client-side cache and the average write speed is, surprisingly, slightly higher.
 It is, however, necessary to use relatively large writes.
-Comparing Figure \ref{fig:results_base_8} and Figure \ref{fig:results_8_nobars},  it can be seen that write sizes must be at least 4MB before transfers to the GPU memory outperform writes to the Lustre cache.
+Comparing Figure \ref{fig:results_base_8} and Figure \ref{fig:results_8_nobars},  it can be seen that write sizes must be at least 4MB before transfers to GPU memory outperform writes to the Lustre cache.
 
@@
 The results for the second test series (16 ranks/node) are less clear.
-As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into the GPU's memory perform slightly better than just writing to disk.
+As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into GPU memory perform slightly better than just writing to disk.
 However, write sizes that exceed the capacity of the GPU's memory are actually a little slower than writing straight to the filesystem.
@@
 On the other hand, an application that is running 16 ranks/node is unlikely to leave much system memory free for use by the Lustre client-side cache.
-So, the improvement from caching writes in the GPU's memory might offset the disruption caused by oversubscribing the cores.
+So, the improvement from caching writes in GPU memory might offset the disruption caused by oversubscribing the cores.
 Whether or not this technique provides a net performance improvement in such a configuration will probably depend on the particular application.
@@
 
-In general, the results show that using the GPU's memory to cache filesystem writes provides approximately the same performance as the Lustre client-side cache.
-The difference is that in order to benefit from the Lustre cache, the writes must be spread across a sufficient number of OST's and the memory must actually be available.
-Whether or not these conditions can be met depends on the particular application.
+In general, the results show that using the GPU memory to cache filesystem writes provides approximately the same performance as the Lustre client-side cache.
+The difference is that the Lustre cache uses system memory which may not actually be available.
+Even if it is, the user must ensure that writes are spread across a sufficient number of OST's to make sue of it.
+Neither of these conditions apply when using GPU memory as a cache.
 

### 2015-04-06 cug_paper/04_futurework.tex
Paper: fix a spelling error

@@
 It is well known that some applications do not use all the RAM on the compute node.
-It is debatable whether letting the daemon use the memory is better than simply letting the operating system use it as  Lustre client cache or OS page cacher.
+It is debatable whether letting the daemon use the memory is better than simply letting the operating system use it as  Lustre client cache or OS page cache.
 However, as was discussed in Section \ref{sec:analysis} and Section \ref{sec:conclusion}, the Lustre client-side cache has fairly low limits on the amount of dirty pages it will allow.

### 2015-04-07 cug_paper/03_analysis.tex
Scott's analysis edits

@@
 In order to evaluate this software, the authors ran two series of tests: one that ran the application with 8 ranks per node and another that ran with 16 ranks per node.
-Each series included tests of the daemon using 1, 2 \& 4 background write threads.
+Each series included tests of the daemon using 1, 2, and 4 background write threads.
 Both series also tested the write performance just using standard \texttt{write()} calls in order to get some baseline data for comparison.
@@
 For the first test series, the application was configured for 8 ranks per node.
-As mentioned in the introduction, 'real' applications will often run on Titan using only 8 ranks per node because the AMD processor in the compute nodes only has 8 floating point units.
+As mentioned in the introduction, `real' applications will often run on Titan using only 8 ranks per node because the AMD processor in the compute nodes only has 8 floating point units.
 From the authors' perspective, this has the advantage of leaving 8 integer-only cores to run the daemon in multi-threaded mode.
@@
 Furthermore, the daemon thread that processed the CCI messages was run in a mode where it continuously polled for new messages.
-This  provided the lowest possible latency for message handling, but at the cost of effectively spinlocking one core.
+This  provided the lowest possible latency for message handling, but at the cost of effectively consuming one core.
 
 For the second test series, the application was configured for 16 ranks per node.
-The daemon was again run with 1, 2 \& 4 write threads.
+The daemon was again run with 1, 2, and 4 write threads.
 For this series, the daemon thread that handled the CCI messages was run in a mode where it would block waiting on a message.
@@
 For both test series, the daemon was configured to write each rank's data to a separate file and the test measured the application's perceived write performance as the write size increased.
-Note the word 'perceived'.
+Note the word `perceived'.
 What was actually measured was how long it took for each rank of the application to copy its data into GPU memory.
@@
 For the second test series, using 16 ranks per node, write sizes up to 256MB would fit.
-For both test series, once the write size exceeded the available GPU memory, the ranks would have to wait until the daemon was able to drain data out of GPU memory and into the filesystem.
+For both test series, once the write size exceeded the available GPU memory, the ranks would have to wait until the daemon was able to drain data out of GPU memory and into the filesystem page cache or over the network.
 
@@
 Figure \ref{fig:results_base_8} shows the results of a baseline test.
-The colored bars show the average values and the error bars display the minimum and maximum values.
+The colored bars show the average throughput for 128 nodes and the error bars display the minimum and maximum values.
 In this test, the application wrote to the filesystem with standard \texttt{write()} calls and the GPU memory was not used at all.\footnote{The daemon process was not even started for this test.}  The results provide some baseline numbers that can be used for comparison with the tests using GPU memory.
@@
 Note the sharp drop in performance between 256MB and 512MB.
-On Titan, the Lustre client is configured to allow a maximum of 64MB per OST and to default to using 4 OST's per file.
-Given the number of OST's available, it's statistically likely that no two output files in this test used the same OST's.
+On Titan, the Lustre client is configured to allow a maximum of 64MB per OST and to default to using 4 OSTs per file.
+Given the number of OSTs available, it is statistically likely that no two output files in this test used the same OSTs.
 In short, write sizes of 256MB or less were cached in system memory using the existing Lustre client cache and the performance of the 512MB, 1GB and 2GB sizes is dominated by the performance of the filesystem.
@@
 Figure \ref{fig:results_8_nobars} shows the remaining results of the first test series.
-As stated earlier, this series was run with 8 ranks per node and with 1, 2 \& 4 write threads in the daemon.
+As stated earlier, this series was run with 8 ranks per node and with 1, 2, and 4 write threads in the daemon.
+Each column is the mean of 64 nodes' perceived per-process throughput.
 The graph shows a number of interesting features.
@@
 
-Concentrating on the single thread performance, it's clear that the application benefits from using GPU memory out to the 512MB write size.
+Concentrating on the single thread performance, it is clear that the application benefits from using GPU memory out to the 512MB write size.
 This makes sense since eight ranks each writing 512MB is a total of 4GB and that will fit into the available GPU memory.
 Even at 1GB, the application sees somewhat improved performance because there is enough GPU memory to hold significant fraction of the data to be written.
-It's not until the 2GB write size that the write performance is dominated by the filesystem's throughput.
+It is not until the 2GB write size that the write performance is dominated by the filesystem's throughput.
 
@@
 This is expected, since there are twice as many ranks writing.
+Again, the baseline test uses 128 nodes while each of the daemon with 1, 2, and 4 write threads present the mean of 64 nodes each.
 
 Figure \ref{fig:results_16_nobars} shows that again, the per-rank bandwidth is about half that shown in Figure \ref{fig:results_8_nobars} because there are twice as many ranks.
-It's also clear that the application ranks see good write performance up to the 256MB write size, which is the largest size that will entirely fit into GPU memory.
-It's also clear that, as in the first test series, writes must be at least 4MB in order to get reasonable performance copying the data into GPU memory.
+It is also clear that the application ranks see good write performance up to the 256MB write size, which is the largest size that will entirely fit into GPU memory.
+It is also clear that, as in the first test series, writes must be at least 4MB in order to get reasonable performance copying the data into GPU memory.
 

### 2015-04-07 cug_paper/05_conclusion.tex
Scott's conclusion edits

@@
 As with the first test series, write sizes need to be at least 4MB Looking closely at Figure \ref{fig:results_base_16} and Figure \ref{fig:results_16_nobars}, it appears that write sizes that can fit into GPU memory perform slightly better than just writing to disk.
-However, write sizes that exceed the capacity of the GPU's memory are actually a little slower than writing straight to the filesystem.
+However, write sizes that exceed the capacity of the GPUs memory are actually a little slower than writing straight to the filesystem.
 
@@
 First, Since the test application did no actual computation, there were plenty of cores available for the daemon to execute on.
-In a real HPC application, this wouldn't be the case and the application's performance would almost certainly be negatively affected by the context switches.
+In a real HPC application, this would not be the case and the application's performance would almost certainly be negatively affected by the context switches.
 On the other hand, an application that is running 16 ranks/node is unlikely to leave much system memory free for use by the Lustre client-side cache.
@@
 It would also be possible to use the  core specialization option to aprun to schedule OS tasks on that core.
-Compared to an application running lots of MPI collectives, the daemon isn't particularly sensitive to short interrupts like that.
-This configuration hasn't been tested and the authors are unaware of any application that routinely runs with 15 ranks/node.
+Compared to an application running lots of MPI collectives, the daemon is not particularly sensitive to short interrupts like that.
+This configuration has not been tested and the authors are unaware of any application that routinely runs with 15 ranks/node.
 
@@
 The difference is that the Lustre cache uses system memory which may not actually be available.
-Even if it is, the user must ensure that writes are spread across a sufficient number of OST's to make sue of it.
+Even if it is, the user must ensure that writes are spread across a sufficient number of OSTs to make use of it.
 Neither of these conditions apply when using GPU memory as a cache.

### 2015-04-07 cug_paper/01_introduction.tex
Paper: Minor edits

@@
 That is, there will be long periods with no file output while the application is in a compute phase, followed by a period of intense output and no computation when the application writes its checkpoint or output files.
-At ORNL, the filesystem is a center-wide resource shared with other systems including a Cray XC-30, analysis clusters, and WAN-connected data mover nodes.
-Each of these competing systems place a varying load on the filesystem such that Titan's users' write performance will vary dramatically within a job and between jobs.
+At ORNL, the filesystem is a center-wide resource shared with multiple systems including a Cray XK7, a Cray XC30, analysis clusters, and WAN-connected data mover nodes.
+Each of these competing systems place a varying load on the filesystem and thus the performance that individual users see can vary dramatically within a job and between jobs.
 

### 2015-04-07 cug_paper/02_design.tex
Paper:  More edits & proofreading.  Last of the /todo boxes removed.
Paper: Minor edits

@@
 \label{subsec:overview}
-To test this idea, we created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
+To test this idea, the authors created a synthetic benchmark application that consists of two executables: the main application which uses MPI and simulates the I/O patterns of a typical HPC application and a daemon process that is started on each compute node.
 The daemon has two tasks: manage the GPU memory and copy data from GPU memory out to a file.
@@
 It then notifies the background write thread(s) that the block is now ready to be written to the filesystem.
-All daemon threads are pinned to unused integer cores.
 
@@
 Since software described in this paper requires access to the GPU from multiple processes, users must switch to the appropriate compute mode.
-This is done by passing the flag ``l feature=gpudefault'' to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.
+This is done by passing the flag ``-l feature=gpudefault'' to qsub.\footnote{It may seem strange to have to explicitly switch to a mode called `gpudefault'.
 The naming scheme comes from the CUDA documentation.
@@
 Titan is configured differently, however.
-Thus the need to explicitly tell qsub to switch the GPU's to the `default' mode.}
+Thus the need to explicitly tell qsub to switch the GPUs to the `default' mode.}
 
@@
 Normally, executables are started on the compute nodes by \emph{aprun}, Cray's replacement for mpirun.
-by default, aprun is designed to start a single executable in single-program-multiple-data (SPMD) mode, not to start two different executables.
+By default, aprun is designed to start a single executable in single-program-multiple-data (SPMD) mode, not to start two different executables.
 Aprun does support multiple-program-multiple-data (MPMD) mode, but it will not start two different executables on the same node (i.e.

### 2015-04-07 cug_paper/03_analysis.tex
Paper: Minor edits

@@
 This added some latency to the message processing, but left that core free to perform useful work when there were no messages to process.
-Also for this series, no core pinning was used on the daemon.\footnote{In practical terms, a user would probably not want to use multiple write threads on the daemon if his/her applications was running 16 ranks/node since it would oversubscribe the cores.
+Also for this series, no core pinning was used on the daemon.\footnote{In practical terms, a user would probably not want to use multiple write threads on the daemon if his/her application was running 16 ranks/node since it would oversubscribe the cores.
 The authors tested the daemon with 2 and 4 threads partially out of curiosity and also to keep the two test series as similar to each other as possible.}
